---
title: "Manuscript"
output: html_document
csl: "pnas.csl"
bibliography: pnas.bib
indent: TRUE
---
```{r manuscript setup, include=FALSE}
#devtools::install_github("an-bui/calecopal")
#devtools::install_github("crsh/citr")
#devtools::install_github("benmarwick/wordcountaddin",  type = "source", dependencies = TRUE)

knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(devtools, tidyverse, openxlsx, ggpubr, here, scales,
               brms, calecopal, citr, knitr, wordcountaddin)

load('pnas.RData')
source(here("R/project_functions.R"))
```

<!-- PNAS is format-neutral at initial submission, which means that manuscripts do not need to be formatted according to specific journal guidelines to be considered for review. We do, however, require the following information in order to evaluate your manuscript: -->

<!-- a manuscript file (in any format) including the following: -->
<!-- title page (title, author list, classification, keywords) -->
<!-- abstract -->
<!-- significance statement -->
<!-- main text -->
<!-- references -->
<!-- figures or tables with appropriate legends (may be uploaded separately) -->
<!-- SI files (may be uploaded separately) -->
<!-- contact and competing interest information for all authors -->
<!-- data sharing plans (for all data, documentation, and code used in analysis) -->
<!-- funding information and whether an open access license has been selected -->
<!-- list of appropriate Editorial Board members, NAS members, and qualified reviewers (minimum of three each) who are experts in the paper's scientific area. (A brief justification for suggested reviewers is welcome, particularly for interdisciplinary papers) -->

<!-- Title page -->
<!-- Abstract -->
<!-- Explain to the general reader the major contributions of the article -->
<!-- Include no more than 250 words -->
<!-- Cite all references in the abstract in full within the abstract itself AND in the text -->
<!-- Significance statement (Direct and Contributed Submissions only) -->
<!-- Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty -->
<!-- Include no more than 120 words -->
<!-- Main text -->
<!-- Introduction -->
<!-- Results -->
<!-- Discussion -->
<!-- Materials and methods (describe procedures in sufficient detail so that the work can be repeated) -->
<!-- Acknowledgments and funding sources -->
<!-- Spell out all abbreviations -->
<!-- Use FundRef to identify the standard name for any funders -->
<!-- Do not include dedications -->
<!-- References -->
<!-- Figure legends -->

# Title Page

## Title: Female reproductive state and ecological conditions impact the magnitude of sex differences in human immune status across the lifespan

## Authors
<!-- Author affiliations -->
<!-- Include, in this order, department/laboratory/section/division, institution, city, state with ZIP code (for US institutions) or country with postal code (for non-US institutions). -->
<!-- Use superscripts to match authors with institutions. Multiple affiliations are allowed. Affiliations where the work was done should be listed. If an author's affiliation has changed since the work was done, the current affiliation should also be listed and noted as such. -->
<!-- PNAS strongly encourages authors to supply an ORCID identifier for each author. Do not include ORCIDs in the manuscript file; individual authors must link their ORCID account to their PNAS account at www.pnascentral.org. For proper authentication, authors must provide their ORCID at submission and are not permitted to add ORCIDs on proofs. -->

Carmen Hove\*^1,2^, Michael D Gurven^1^, Benjamin C Trumble^3^, Jonathan Stieglitz^4^, Daniel Eid Rodriguez^5^, Ivan Maldonado Suarez^6^, Hillard Kaplan^7^, Aaron D Blackwell^+8^, Amy M Boddy^+1^

^1^ Department of Anthropology, University of California Santa Barbara, Santa Barbara, California, 93106 \
^2^ Reda Analytics, Seattle, Washington, 98119 \
^3^ Center for Evolution and Medicine, School of Human Evolution and Social Change, Institute of Human Origins, Arizona State University, Tempe, Arizona, 85287 \
^4^ Institute for Advanced Study in Toulouse, Toulouse School of Economics, Toulouse, France, 31015 \
^5^ Department of Medicine, Universidad de San Simón, Cochabamba, Bolivia \
^6^ Tsimane Health and Life History Project, San Borja, Bolivia \
^7^ Economic Science Institute, Chapman University, Orange, California, 92866 \
^8^ Department of Anthropology, Washington State University, Pullman, Washington, 99164 \

\*Corresponding Author \
^+^Shared Senior \

Carmen Hove \
Email: carmen@redaanalytics.com \
Phone: (360) 621-3508 \

<!-- Aaron Blackwell \ -->
<!-- Email: aaron.blackwell@wsu.edu \ -->
<!-- Phone: (805) 253-2796 -->

<!-- Carmen Hove* \ -->
<!-- ORCID: https://orcid.org/0000-0002-8626-321X \ -->

<!-- Michael Gurven \ -->
<!-- ORCID: https://orcid.org/0000-0002-5661-527X \ -->
<!-- Email: gurven@anth.ucsb.edu \ -->

<!-- Ben Trumble \ -->
<!-- ORCID: https://orcid.org/0000-0003-3201-0628 \ -->
<!-- Email: trumble@asu.edu \ -->

<!-- Jonathan Stieglitz \ -->
<!-- ORCID: https://orcid.org/0000-0001-5985-9643 \ -->
<!-- Email: jonathan.stieglitz@gmail.com \ -->

<!-- Daniel Eid Rodriguez -->
<!-- ORCID: https://orcid.org/0000-0001-8145-0967 -->

<!-- Ivan Maldonado Suarez -->
<!-- ORCID: https://orcid.org/0009-0003-7345-6519 -->

<!-- Hillard Kaplan \ -->
<!-- ORCID: https://orcid.org/0000-0002-7398-7358 \ -->
<!-- Email: hkaplan@chapman.edu \ -->

<!-- Amy Boddy \ -->
<!-- ORCID: https://orcid.org/0000-0002-7990-2415 \ -->
<!-- Email: boddy@anth.ucsb.edu \ -->

<!-- Aaron Blackwell* \ -->
<!-- ORCID: https://orcid.org/0000-0002-5871-9865 \ -->
<!-- Affiliation: Washington State University -->

## Classification
<!-- Select both a major and minor classification category. Dual classifications are permitted between major categories. -->

Major: Biological Sciences \
Minor: Immunology and Inflammation

## Keywords
<!-- Include at least three keywords at submission. -->

\newpage

## Abstract
<!-- Explain to the general reader the major contributions of the article. Include no more than 250 words. -->

In industrialized societies, women benefit from lower infectious disease burden than men but suffer disproportionately from autoimmune disease. Considering the role that hormones play in both female reproduction and the development of sex differences in immune function, sexual dimorphism in immune status may vary by both current female reproductive state and female reproductive history. Furthermore, socio-ecological conditions common to industrialized populations (e.g., energetic excess, low pathogen exposure) may exacerbate evolved sex differences in immunity. Here, we model age-specific sexual dimorphism, stratified by female reproductive state and controlling for parity, in leukocyte differential and neutrophil-to-lymphocyte ratio (NLR) among the Tsimane, a non-industrialized population, and the USA. We find that sex differences in most immune measures are greatest between pregnant women and their male peers. We find non-linear effects of age and negative effects of parity on neutrophil count and NLR among pregnant women in the USA but not among the Tsimane. Likewise, we find that menopause corresponds to a reversal of sex biases in most immune measures in the USA but not among the Tsimane. These results suggest that while pregnancy may be a universal driver of sex differences in immune status, changes in reproductive schedules (later age at first birth and reduced parity) may alter the immunological legacy of pregnancy. Finally, the population differences we observe in this study indicate that oversampling of post-industrial populations has skewed our current understanding of "normal" sexual dimorphism in immune status and disease risk. 

## Significance Statement
<!-- Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty. Include no more than 120 words. -->

Current research indicates that women generally mount greater immune defenses against infectious agents but suffer increased risk of autoimmune disease. Given that most research on sex differences in immune function comes from industrialized urban populations and often excludes women in different reproductive phases (e.g., pregnancy, menopause), it is unclear how the magnitude and direction of sex bias in immune status are influenced by current and/or prior female reproductive effort or broader ecological conditions. Here we show that maximum sexual dimorphism in immune measures across the lifespan is generally lower among the Tsimane (a natural-fertility non-industrialized population) compared to the USA, due to population-specific effects of age and parity on female immune status during pregnancy and following menopause. 

\newpage

# Main Text

## Introduction

In post-industrial human populations, women benefit from reduced infectious disease burden and lower risk of non-reproductive cancers relative to men but make up approximately 80% of autoimmune disease diagnoses [@Whitacre2001a] and suffer disproportionately from allergy and atopy [@Laffont2017]. These sex biases in disease risk are reflected in sex differences across immune biomarkers [@Klein2016]. Women generally possess more neutrophils [@bain_england75; @casimir_etal10; @chen_etal16a; @li_etal20], granulocytes that orchestrate first-line defenses against pathogens, prime antigen-specific immunity, and induce potentially deleterious inflammatory cascades [@chen_etal14; @mayadas_etal14; @polak_etal19]. Women also have higher B cell and CD4^+^ T cell counts [@abdullah_etal12; @uppal_etal03]. Conversely, total lymphocyte count is often higher in males [@abdullah_etal12], reflecting higher numbers of regulatory T [@afshan_etal12] and natural killer cells [@abdullah_etal12]. Consequently, women frequently possess a comparatively higher neutrophil-to-lymphocyte ratio (NLR). NLR is a robust indicator of immune homeostasis and systemic inflammation [@kourilovitch_galarza-maldonado23] strongly associated with the presence and severity of numerous autoimmune disorders that disproportionately affect women (e.g., rheumatoid arthritis) [@zhou_etal23; @damico_etal19; @hasselbalch_etal18; @mihai_etal22; @wang_etal20]. While there is ongoing debate regarding the ultimate selective pressures responsible for such sex differences in immune function [@McKean2005; @metcalf_graham18a; @fink_klein18; @mitchell_etal22; @Natri2019], sex-specific hormone production [@kocar_etal01; @gayen_etal16; @Taneja2018; @furman_etal14] is an evolutionarily conserved proximate mechanism by which females mount stronger cellular and humoral responses to pathogens [@Fish2008b; @Takahashi2020], vaccination [@klein_etal10; @Mitchell1999; @Voigt2019], and auto-antigens.  

<!-- Proposed selective pressures responsible for producing such sex differences in immune function include mate choice and the corresponding development and maintenance of "costly" secondary male sex characteristics [@McKean2005], sex-specific reproductive schedules [@metcalf_graham18a], the need for increased transfer of antibodies to offspring during and after pregnancy [@fink_klein18], vertical transmission of pathogens from mother to offspring during pregnancy [@mitchell_etal22], and "the presence of an invasive placenta and an immunologically challenging pregnancy" (Pregnancy Compensation Hypothesis) [@Natri2019]. -->

<!-- incomplete X-inactivation of immune-related genes during female development [@gayen_etal16] and  -->

Because female hormone production varies by reproductive state (premenarchal, regularly menstruating, pregnant, postpartum, and postmenopausal), the direction and magnitude of sex biases in immune function are also expected to vary across these phases. However, the impact of such *within-sex* variability on overall sex differences in immune status is not well documented. Likewise, the effect of cumulative reproductive effort (e.g., total number of pregnancies) on female immune function and corresponding sexual dimorphism in immune status is under-explored. The Pregnancy Compensation Hypothesis (PCH) suggests that sex differences in immune function are amplified in low-fertility populations compared to those in the evolutionary past. During pregnancy, the maternal immune system undergoes a compensatory shift, modulated by hormones such as progesterone and estrogens, to support fetal tolerance while maintaining host defense. These evolved adaptations are thought to recalibrate the immune system in ways that persist beyond pregnancy. However, for women with low parity, which is common in industrialized societies, this compensation may become dysregulated, leading to heightened sex differences in immune function and increased susceptibility to immune-related disorders [@Natri2019]. This hypothesis is supported by evidence that certain autoimmune diseases go into remission during pregnancy [@adamswaldorf_nelson08a] while infectious disease risk is temporarily increased [@robinson_klein12], mimicking a more "male-typical" risk profile. On the other hand, many autoimmune diseases flare and/or emerge after pregnancy [@khashan_etal11], suggesting that the immunological legacy of gestation is more complicated. To date, however, few studies have investigated the effects of parity on sexual dimorphism in immune function. Furthermore, most research on sex differences in immune function has been conducted in post-industrial or rapidly industrializing populations, where evolutionarily novel conditions, such as reduced microbial exposure and lower energetic demands, may exaggerate underlying sex differences in immunity. These environments limit opportunities for immune system calibration during development and may heighten sensitivity to sex hormones.

In this study, we investigate overall sex differences in immune function as well as variation in immune function among females in different reproductive states across two ecologically distinct populations: a heavily industrialized U.S. population and the Tsimane, a natural-fertility subsistence population. We infer differences in immune activity by examining the distribution of immune cell counts across the lifespan in each population. To test the Pregnancy Compensation Hypothesis [@Natri2019], we also estimate and compare population-specific effects of parity on immune markers, stratified by female reproductive state. As indicators of immune status, we focus on white blood cell differentials and neutrophil-to-lymphocyte ratios, biomarkers that reflect broad immunological processes [@heim_etal18; @ishihara_etal06] and have well-established diagnostic and prognostic relevance for numerous health outcomes [@al-gwaiz_babay07; @kourilovitch_galarza-maldonado23; @lowsby_etal15] (Table 1). To fully understand how sex differences in immunity arise and vary, it is important to consider how female reproductive state, reproductive history, and ecological context impacts immune function across the lifespan. These factors influence both hormone production and immunological activity and may help explain why sex biases in immune profiles, and their associated health outcomes, can differ across populations and life stages.

### Female reproduction requires substantial hormonal and immunological shifts
 
While aging impacts hormone production [@handelsman_etal16; @decaroli_rochira17] and immune senescence [@marquez_etal20] in both sexes, female reproduction requires additional hormonal changes and presents unique immunological challenges. During pregnancy, production of estradiol, estrone, testosterone, and progesterone are significantly elevated above baseline [@oleary_etal91; @schock_etal16]. These hormonal shifts support fetal development [@magon_kumar12] and fetal tolerance, a phenomenon in which the maternal immune system must tolerate fetal antigens while maintaining essential immune defenses [@schumacher_etal14a; @soma-pillay_etal16]. Following delivery, ovarian function shifts again: estrogen and progesterone production are suppressed by the antagonistic effects of prolactin, a hormone which remains elevated in the presence of regular lactation [@neville_neifert83]. There is some evidence that pregnancy produces lasting alterations in the production of certain hormones, including reduced prolactin secretion [@musey_etal87a] and elevated estriol [@musey_etal87], but how these long-term alterations impact immune function is not well understood. Finally, menopause marks the end of the female reproductive lifespan, with the ovaries switching to low production of estrogens and progesterone and continued production of androgens [@burger02]. Although males do not experience an exact physiological equivalent to menopause, advancing age in men is also marked by significant declines in sex hormone production [@decaroli_rochira17] and overall fertility [@khandwala_etal18; @pino_etal20].  

The dynamic nature of ovarian function over the life course suggests that hormone-mediated sex biases in immune function and disease risk should be most pronounced between puberty and menopause. Given the steep increase in estrogens and progesterone observed during gestation and evidence that systemic inflammation *increases* during pregnancy [@Hove2020], sex bias in immune measures may be further accentuated by pregnancy. In line with these predictions, autoimmune diseases that most disproportionately affect women (e.g., primary Sjögren's syndrome, systemic lupus erythematosus, Hashimoto thyroiditis) are most commonly diagnosed among women between 20 and 50 years of age [@garcia-carrasco_etal02; @ohta_etal13; @mincer_jialal24; @moinzadeh_etal20; @aarli99] and often emerge or worsen after pregnancy [@piccinni_etal16]. On the other hand, the immunological shifts induced by fetal and placental cues during pregnancy may *reduce* sex differences in immune measures [@Natri2019]. Regular exclusion of pregnant women and failure to consider the effects of parity preclude a clear picture of how age, current female reproductive status, and female reproductive history combine to impact sex differences in immune status.

### Hormone production and immunological development are sensitive to ecological inputs

Socio-ecological conditions that increase overall energy balance (e.g., caloric excess, increased sedentism) may magnify hormone-mediated sex differences in immune function via disproportionately large effects on female ovarian function. Because female reproduction across mammalian species requires significantly greater energetic investment than male reproduction, female ovarian function is particularly responsive to energy balance. While reduced energetic availability suppresses baseline ovarian function, resulting in lower progesterone production across the ovarian cycle [@Jasienska1998a; @nunez-delamora_etal07; @vitzthum_etal04a], positive energy balance and reduced metabolic load are linked to earlier age at menarche [@thomas_etal01], elevated adult levels of estradiol [@emaus_etal08] and progesterone [@nunez-delamora_etal07], higher progesterone levels during the peri-ovulatory and peri-implantation period [@vitzthum_etal04a], and shorter duration of lactational amenorrhea among breastfeeding women [@valeggia_ellison09]. While moderate energetic availability is associated with enhanced fecundity [@sadhir_pontzer23] and greater immune competence [@childs_etal19], caloric excess contributes to chronic inflammation and greater risk of hypersensitivity to a broad range of antigens [@macsali_etal09]. Furthermore, exposure to *exogenous* female sex hormones via regular use of oral contraceptives, an evolutionarily novel feature of industrialized societies, has also been linked to immunological hypersensitivity to a variety of antigens - including steroid hormones themselves [@untersmayr_etal17]. 

Lack of exposure to certain pathogens during development may also increase sexual dimorphism in immune function, not just by increasing the amount of energy allocated to sex-specific hormone production, but by reducing the number of opportunities for immunological calibration. In the United States, for example, reductions in infectious disease burden [@armstrong_etal99; @roush07] have increased life expectancy [@cutler_miller05] but are linked to the rise of chronic inflammatory disorders characterized by hypersensitivity to non-pathogenic antigens (e.g., atopy, autoimmune disease) [@bloomfield_etal16]. One explanation for this relationship is that exposure to pathogens during development primes the immune system to differentiate between pathogenic and non-pathogen antigens and, in certain cases, induce immunological tolerance. People living in a non-industrialized context characterized by elevated pathogen load exhibit higher baseline levels of most immune markers [@Blackwell2011c; @Blackwell2016] but seemingly low incidence of allergy or autoimmune disease [@gurven_etal07; @Gurven2008a; @kaplan_etal17; @wu_etal23]. Lack of calibrating opportunities via pathogen exposure may have particular relevance during pregnancy, when the maternal immune system must induce tolerance of fetal antigens. Current evidence indicates that non-industrialized populations experiencing high pathogen exposure exhibit less immunological activation (e.g., lower peak in neutrophil count and C-reactive protein) during pregnancy [@Hove2020], which may reflect both lower hormone production and reduced immunological sensitivity to fetal antigens during gestation. 

Taken together, these patterns suggest that evolutionarily novel environmental conditions common to industrialized populations (e.g., low energetic throughput, microbial deprivation) may exacerbate evolved sex differences in immunity by reducing opportunities for immunological calibration during development and increasing exposure and/or sensitivity to sex hormones - especially among females. At time of writing, however, these ideas remain largely unexplored and therefore current definitions of "normal" sexual dimorphism in immune status rest primarily on studies conducted within industrialized/industrializing societies (e.g., China, the USA, India).

## Objectives and predictions

We utilize data from the Tsimane, a natural-fertility subsistence population inhabiting the Amazonian River basin, and a representative sample from the United States to estimate the age-dependent effects of sex, stratified by female reproductive phase (premenarchal, regularly cycling, pregnant, 0-12 months postpartum, and postmenopausal), on white blood cell counts and neutrophil-to-lymphocyte ratio using generalized additive models. Among postmenarchal females, we also estimate the population-specific effects of parity on immune measures, stratified by female reproductive state.

Within both populations, we predict that sex biases in immune status will be most pronounced between postmenarchal and premenopausal females and their age-matched male peers, with pregnancy corresponding to greater sexual dimorphism. Given the reversal of pregnancy-induced hormonal patterns and suppression of ovulation following delivery, we expect that postpartum females will exhibit less divergent immune profiles from their male counterparts. Likewise, we hypothesize that sex differences in immune status will be attenuated or reversed among postmenopausal females and their male peers, due to the combined effects of aging in both sexes and menopause among females. Regardless of female reproductive state, we expect that sex biases in immune measures will be attenuated among the Tsimane compared to the USA. Finally, we test the Pregnancy Compensation Hypothesis' core prediction that reduced parity will be associated with a larger degree of sex bias in industrialized populations [@Natri2019]. 

<!-- Here, we utilize generalized additive models to estimate the age-dependent effects of sex, stratified by female reproductive phase (premenarchal, regularly cycling, pregnant, 0-12 months postpartum, and postmenopausal), on white blood cell counts and neutrophil-to-lymphocyte ratio among the Tsimane, a natural-fertility subsistence population inhabiting the Amazonian River basin, and a representative sample from the United States. We predict sexual dimorphism in leukocyte differential and NLR will be greatest among reproductive-aged individuals. Since estradiol and progesterone levels are elevated above baseline during pregnancy, immune changes during gestation may drive a large proportion of sexual dimorphism within each population. Given the reversal of pregnancy-induced hormonal patterns, suppression of ovulation, and the removal of the need for fetal tolerance, postpartum females may exhibit less divergent immune profiles from their male counterparts. Given the drop-off in female hormone production after menopause, we expect that sexual dimorphism in immune status will be reduced among postmenopausal women and age-matched men. We also estimate the effects of parity on immune status among all postmenarchal females to test the Pregnancy Compensation Hypothesis' prediction that reduced parity exacerbates sexual dimorphism [@Natri2019].  -->

## Results

### Female reproductive status has age-dependent and population-specific effects on the direction and magnitude of sex differences in immune markers. 

Among males in both populations, age is associated with strong non-linear and population-specific effects on all immune measures, especially within the first two decades of life (Figure 1, Table S4, Table S5). Among females, the effects of age on immune measures are mediated by current reproductive state and differ based on population and immune measure (Figure 1, Table S4, Table S5). 

#### Among premenarchal females and their age-matched male peers, certain ages are marked by a male bias in eosinophil and monocyte acounts in the USA versus a female bias in eosinophil and monocyte count among the Tsimane. 

In the USA, premenarchal females have lower eosinophil counts than males between the ages of `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Max"]`, lower monocyte counts between the ages of `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Max"]`, and slightly higher total lymphocyte counts from ages `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Max"]` and again at age `r get.values(table_2, "NHANES","LYM", "Pre-Menst", 2)[,"Min"]`. We find no differences in neutrophil count, total white blood cell count, and NLR between premenarchal females and their male peers within the USA, regardless of age (Figure 1A, Figure 1C, Table 2). Among the Tsimane, premenarchal females have slightly *lower* total lymphocyte counts, but only from ages `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Max"]`. Likewise, premenarchal females possess slightly *higher* monocyte counts than males, but only from ages `r get.values(table_2, "THLHP","MON", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "THLHP","MON", "Pre-Menst")[,"Max"]`. 

#### Sex differences between regularly cycling females and their male counterparts is limited in both populations, with the exception of enduring male bias in eosinophil and monocyte counts among individuals in the USA. 

Similar to premenarchal females, regularly cycling females in the USA have lower eosinophil counts than males at all ages and lower monocyte counts between ages `r get.values(table_2, "NHANES","MON", "Cycling")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Cycling")[,"Max"]` (Figure 1A). As shown in Table 2, regularly cycling females in the USA have higher neutrophil and total white blood cell counts than males at younger and later ages, but most ages are characterized by an *absence* of robust sex biases in these immune measures. Likewise, there is a female bias in NLR between regularly cycling females in the USA and their male counterparts, but only from ages `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Min"]` to `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Max"]`. Among individuals at younger ages, regularly cycling Tsimane women have lower neutrophil, total lymphocyte, eosinophil, and total white blood cell counts and higher NLR than males - but most ages are characterized by an *absence* of robust sex biases in these immune measures (Figure 1B, Table 2). 

#### In both populations, pregnancy is marked by a male bias in total lymphocyte and eosinophil counts. In the USA, pregnancy is also marked by a robust age-dependent female bias in neutrophil count, neutrophil-to-lymphocyte ratio, and total white blood cell count. 

In contrast to regularly cycling females, pregnant females in the USA have higher neutrophil count, total white blood cell count, and NLR than their male peers for the majority of ages represented, with highly non-linear effects of age producing maximum sex differences at `r get.abs.values(table_s3, "NHANES","NEU","Pregnant")[,"Age"]`, `r get.abs.values(table_s3, "NHANES","WBC","Pregnant")[,"Age"]`, and `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Age"]` years of age, respectively (Figure 1A, Table 2, Table 3). Conversely, pregnant females in the USA have lower total lymphocyte count than males from ages `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Min"]` to `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Max"]` and lower eosinophil counts between ages `r get.values(table_2, "NHANES","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Max"]`. 

Pregnant Tsimane females have lower total lymphocyte count than males from ages `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Min"]` to `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Max"]`, lower eosinophil counts between ages `r get.values(table_2, "THLHP","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Max"]`, and lower total white blood cell count from ages `r get.values(table_2, "THLHP","WBC", "Pregnant")[,"Min"]` to `r get.values(table_2, "THLHP","WBC", "Pregnant")[,"Max"]`. 

#### Among postpartum women, there is a male bias in monocyte counts in the USA and a female bias in monocyte counts among the Tsimane. 

Postpartum females in the USA have lower monocyte counts than males from ages `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Max"]` but otherwise exhibit no differences from their male counterparts. Postpartum females in the Tsimane have *higher* monocyte counts than males from ages `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Min"]` and `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Max"]` but otherwise exhibit no differences from their male counterparts. 

#### Sex differences 

Finally, depending on age, postmenopausal females in the USA have *lower* neutrophil, eosinophil, and monocyte counts, *higher* total lymphocyte count, and lower neutrophil-to-lymphocyte ratio than males (Figure 1A, Table 2). Depending on age, postmenopausal Tsimane women have *lower* NLR and eosinophil counts than males (Figure 1, Table 2).

<!-- #### The direction and magnitude of sexual dimorphism in immune status varies between the USA and the Tsimane -->
<!-- The results described above create age-dependent differences in both the direction and magnitude of sexual dimorphism in immune cell counts and NLR between the Tsimane and the USA. Among premenarchal females and their male peers, there is a female bias in total lymphocyte count and male bias in eosinophil and monocyte counts in the USA but only a short-lived *male* bias in total lymphocyte count among the Tsimane. Among regularly cycling females and their male peers, there is a female bias in neutrophil count, NLR, and total white blood cell count and sustained male bias in eosinophil and monocyte counts in the USA versus a *male* bias in neutrophil, total lymphocyte, and white blood cell count among the Tsimane. Among pregnant females and their male counterparts, we find a female bias in NLR and a male bias in total lymphocyte and eosinophil counts in both populations. In the USA, however, female bias in NLR is further elevated by a corresponding female bias in neutrophil count that we do not find among the Tsimane. Among postpartum females and their male peers, we find an age-dependent male bias in monocyte count in the USA and an age-dependent female bias in monocyte count among the Tsimane. Among postmenopausal females and their male peers, there is a female bias in total lymphocyte count and male bias in neutrophil count, eosinophil count, monocyte count, and NLR in the USA. Among the Tsimane, the only sex bias we observe among postmenopausal women and age-matched men is a slight male bias in NLR among elderly individuals.   -->

### Higher parity corresponds with reduced sex bias in neutrophil count and NLR - but only among pregnant women in the USA

We do not find any robust effects of parity on immune markers among regularly cycling, pregnant, or postpartum Tsimane females. Likewise, we do not find strong effects of parity on immune measures among regularly cycling or postpartum females in the USA (Figure 2, Table S4, Table S5). We do, however, find strong negative effects of parity on neutrophil count `r get.statistics(model_summaries, "USA","NEU","s(Parity):Pregnant")`, monocyte count `r get.statistics(model_summaries, "USA","MON","s(Parity):Pregnant")`, and NLR `r get.statistics(model_summaries, "USA","NLR","s(Parity):Pregnant")` among pregnant females in the USA (Figure 2, Table S5). Consequently, high-parity pregnant women in the USA exhibit immune profiles that are more similar to age-matched men compared to primiparous women. Controlling for BMI and age, the estimated neutrophil count, monocyte count, and NLR for a *currently pregnant* nulliparous woman in the USA is `r get.ci(parity_comps, "NEU", "USA", 0, "Pregnant")`, `r get.ci(parity_comps, "MON", "USA", 0, "Pregnant")`, and `r get.ci(parity_comps, "NLR", "USA", 0, "Pregnant")`, respectively. A pregnant woman with 4 prior live births has a `r get.percent(pred_list[["USA.NEU"]], 0, 4, "Pregnant","Pregnant")` lower neutrophil count, a `r get.percent(pred_list[["USA.MON"]], 0, 4, "Pregnant","Pregnant")` lower monocyte count, and `r get.percent(pred_list[["USA.NLR"]], 0, 4, "Pregnant","Pregnant")` lower NLR compared to a currently pregnant nulliparous woman of the same age. 

<!-- For a pregnant woman with 4 prior live births, estimated neutrophil count, monocyte count, and NLR is `r get.ci(parity_comps, "NEU", "USA", 4, "Pregnant")`, `r get.ci(parity_comps, "MON", "USA", 4, "Pregnant")`, and `r get.ci(parity_comps, "NLR", "USA", 4, "Pregnant")`, respectively.  -->

### Post menopausal

Among postmenopausal females in both populations, we find a non-linear effect of parity on NLR (Figure 2, Table S4, Table S5). Among postmenopausal Tsimane females, we observe a negative effect of parity on eosinophil count `r get.statistics(model_summaries, "Tsimane","EOS","s(Parity):Postmenopausal")`. Controlling for BMI and age, the estimated eosinophil count for a nulliparous postmenopausal Tsimane woman is `r get.ci(parity_comps, "EOS", "Tsimane", 0, "Postmenopausal")` and `r get.ci(parity_comps, "EOS", "Tsimane", 4, "Postmenopausal")` for a postmenopausal woman with a history of 4 live births. 

## Discussion

Pregnancy is often described as a period during which autoimmune disease risk is temporarily alleviated while risk of certain infections is higher, resembling a more male-typical risk profile. The results of this study, on the other hand, indicate that pregnancy is a primary *driver* of sex differences in immune cell counts. In both populations, sexual dimorphism in neutrophil-to-lymphocyte ratio (a simple yet robust indicator of immunological homeostasis) is most pronounced between pregnant women and age-matched men, with pregnant women exhibiting substantially higher NLR. Among pregnant women in the USA, we find robust non-linear effects of age and strong negative effects of parity on neutrophil count and NLR. In the USA, estimated neutrophil count and NLR are therefore highest among primiparous pregnant females around 29 years of age, with NLR values exceeding pre-established thresholds of  "mild to moderate inflammation" [@zahorec21]. Supporting some of the predictions in the Pregnancy Compensation Hypothesis, we do not observe these same effects of age or parity among pregnant Tsimane women and therefore maximum sex differences (as well as within-sex variation) in neutrophil count and NLR among the Tsimane are much smaller than those observed in the USA. However, the mechanisms driving these sex-differences in the US are more complicated and these results suggest a refinement of the Pregnancy Compensation Hypothesis, including considerations of age at first birth. At a mechanistic level, high NLR among women in the USA who become pregnant for the first time in their late-twenties to early-thirties may be due to elevated baseline estrogen levels and altered progesterone-to-estradiol ratio. Studies among non-pregnant women in industrialized populations report a non-linear effect of age on estradiol production, with levels peaking around the age of 30 [@lephart_naftolin21]. The absence of strong age effects on immune status among pregnant Tsimane women may reflect lower age-related variability in hormone production due to chronic non-reproductive demands on energy allocation. The absence of strong parity effects on immune measures among the high fertility Tsimane suggest that the impact of cumulative reproductive output on female immune function may depend on the broader ecological context. 

<!-- Our results suggest that there is something unique about the short and long-term immunological effects of pregnancy among women in industrialized societies.  -->
<!-- The effects of age on hormone production during pregnancy may be further mediated by parity, since pregnancy is associated with potentially long-term reductions in ovarian hormone production [@arslan_etal06; @barrett_etal14].  -->

According to the Centers for Disease Control, the average woman in the USA gives birth for the first time at 27 years old and has approximately 1.8 live births over her lifespan. These national trends are remarkably close to the demographic that we find to have acutely elevated NLR during pregnancy. Given the association between high NLR and immunological dysregulation [@damico_etal19; @kim_etal17a; @mihai_etal22; @wang_etal20], it is tempting to speculate that widespread changes in reproductive behavior within industrialized societies (later age at first birth and reduced parity) contribute to excess autoimmune diseases diagnoses among women by altering the immunological legacy of pregnancy. If true, this may explain why many autoimmune diseases flare and/or emerge after pregnancy [@khashan_etal11] and why, when lumped together, women are predominantly diagnosed with these diseases before the average age of menopause [@angum_etal20]. During pregnancy, immediately negative effects of acutely elevated NLR on overall disease risk may be mitigated by the presence of the fetus/placenta and the associated mechanisms that induce fetal tolerance (e.g., regulatory T cell proliferation, regulation of neutrophil phenotypes) [@nadkarni_etal16a; @Erlebacher2013a]. Deleterious effects may then emerge after delivery, when placental cues are removed but offspring cells often remain in the maternal body [@adamswaldorf_nelson08a]. Relatively low rates of extended on-demand breastfeeding after delivery [@flaherman_etal11], an evolutionarily conserved phase of mammalian reproduction often described as the "4th trimester", may further magnify these effects by impeding postpartum immunological recovery [@carmenhove_etal24]. We recommend that future research investigate the more granular effects of time since delivery and breastfeeding behavior on immune function, as this approach may illuminate important effects that we were not able to separate out in this study. 

<!-- Approximately ~84% of females in the USA have at least one child [@martinez_daniels23].  -->

Another way future studies can evaluate the immunological legacy of pregnancy is by comparing short, mid, and long-term health outcomes among women with different reproductive histories. Given that we find comparatively little sexual dimorphism between regularly cycling females and their age-matched male counterparts and only minor effects of parity on immune status among postmenopausal females, it is possible that nulliparous women without any underlying fertility challenges have a *reduced* chance of developing autoimmune disease compared to low-parity women who begin their reproductive careers in their late-twenties or early-thirties. Likewise, our results suggest that risk for autoimmune disease onset following pregnancy may be lower among women who start their reproductive career relatively early. We strongly recommend that future studies consider both age at first birth and total number of pregnancies when investigating sex differences in health outcomes.

While pregnancy is a primary driver of sex differences in immune status in both populations, our results indicate that the transition to menopause is marked by sexual dimorphism in immune status in the USA but not among the Tsimane. In the USA, we find a sustained male bias in neutrophil, eosinophil, and monocyte count and NLR and a female bias in total lymphocyte count among postmenopausal women and age-matched men. In contrast, we find a near-absence of sex differences in immune measures among postmenopausal Tsimane women and their male counterparts. These patterns suggest that the transition to menopause is characterized by a more severe drop-off in ovarian hormone production among females in the USA and/or greater sensitivity of the immune system to the hormonal changes that occur during menopause. In sum, this study shows that women in the USA (especially those who have one or more pregnancies) experience a much higher degree of overall variability in certain immune measures (e.g., neutrophil count and NLR) across the lifespan when compared to Tsimane women, presumably due to greater vacillations in ovarian sex hormone production.

Lastly, we find a consistent male bias in eosinophil and monocyte count across nearly all ages and female reproductive states within the USA but observe no such bias among the Tsimane. These results indicate that monocyte and eosinophil production is less responsive to changes in ovarian sex hormone production across lifespan. Furthermore, the absence of a sustained male bias in eosinophil and monocyte count among the Tsimane suggests that sex differences in these immune measures are not universal. These population differences may arise from the relatively low exposure to infectious diseases in the USA, which could unmask sex-specific effects of gene expression on immune development.

### Additional Considerations
While useful in establishing a general understanding of immune status, variability in cell counts does not perfectly map on to variation in cellular function or underlying gene expression. Future comparative work is needed to investigate the effects of age, sex, female reproductive status, parity, and ecological conditions on a broader range of immune markers (e.g., oxidative bursts and release of extracellular traps by neutrophils). We especially recommend that future research focus on neutrophil subtypes, especially low-density neutrophils (LNDs), as these have been identified as key players in the etiology of autoimmune disease [@ning_etal22]. 

Additionally, the comparisons we draw between the USA and the Tsimane should be interpreted in context. There is significant variation in socio-ecological conditions *within* non-industrialized societies that must be carefully considered, including differences in physical environment (e.g., altitude, seasonality, type of pathogen exposure) and social structure (e.g., population density, matriarchal versus patriarchal customs). For example, the Shuar (another indigenous population living in the Amazon) have distinct pathogen exposure profiles compared to the Tsimane [@Blackwell2011c]. Furthermore, there is considerable socio-ecological variation within industrialized societies (e.g., population density, socio-economic status) that we chose to collapse based on the aims of this particular study. 

### Conclusions
In this study, we show that pregnancy is a primary driver of sex differences in neutrophil and NLR immune cell counts in two socio-ecologically distinct populations. As predicted by the Pregnancy Compensation Hypothesis, we find these sex differences in immune cell counts are more pronounced within the USA when compared to the Tsimane. Considering the impact of age on immune cell counts of pregnant females in the USA, we suggest further refinement of the hypothesis to consider the timing and tempo of reproductive history in females. Given the link between acutely elevated NLR and heightened risk of autoimmunity, we propose that later age at first birth and lower parity may contribute to excess autoimmune diseases diagnoses among women by altering the immunological legacy of pregnancy. While the mechanisms driving these immune cell count differences among populations are unclear, we speculate that females in industrialized populations may experience more dramatic vacillation in hormone production across the lifespan. To address this, we suggest that future research on sex differences in immune function and disease risk collect corresponding data on ovarian hormone levels. Finally, we argue that differences in the magnitude and sometimes direction of sex biases in immune measures between the USA and the Tsimane show that current understanding of these processes is largely limited to the post-industrialized contexts where they have been predominantly studied. We therefore advocate that future research consider the impact of socio-ecological conditions and reproductive history on the physiological and behavioral processes that produce sexual dimorphism in immune function and disease risk.

\newpage

## Methods

### The Tsimane
The Tsimane are a small subsistence population inhabiting the Bolivian Amazonian River basin. Among the Tsimane, chronic exposure to diverse pathogens causes high infectious disease morbidity and mortality across all ages [@gurven_etal19; @kaplan_etal15; @martin_etal13], while incidence of allergies, atopy, autoimmune disease, obesity, and atherosclerosis is low [@gurven_etal07; @Gurven2008a; @kaplan_etal17]. As a result of elevated pathogen burden, Tsimane individuals exhibit high levels of immune activation compared to Western clinical standards [@Blackwell2016; @Blackwell2011c]. Such high investment in immune function results in trade-offs with growth during development [@Foster2005a] and high resting metabolic rate during adulthood [@gurven_etal16], reflecting the substantial energetic demands of coping with pathogenic threats. The Tsimane are also a natural fertility population, with limited access to contraception and breastfeeding alternatives [@trumble_etal23a]. As a result, Tsimane women have an average of 9 live births over the reproductive lifespan [@McAllister2012a] and exhibit nearly ubiquitous rates of on-demand breastfeeding following parturition, with a mean infant weaning age of 27 months [@Martin2016a]. 

### Datasets
We used cross-sectional and longitudinal clinical and demographic data collected by the Tsimane Health and Life History Project (http://tsimane.anth.ucsb.edu/index.html) [@Blackwell2011c; @Blackwell2015c; @Blackwell2016; @gurven_etal17; @martin_etal13] between 2004 and 2014. Approval by the Gran Consejo Tsimane and by institutional review boards at the University of California, Santa Barbara, the University of New Mexico, and the Universidad Mayor San Simon, Cochabamba Bolivia was obtained before any data were collected. Informed consent was provided by participants during a community-wide meeting open to all Tsimane residents and again at the individual level before each medical visit and interview. In the case of minors, parental consent was given before data were collected. Total leukocyte count was obtained via venous blood draws and determined with a QBC Autoread Plus dry hematology system (QBC Diagnostics). Relative fractions of neutrophils, eosinophils, lymphocytes, and monocytes were then measured manually by microscopy with a hemocytometer. Menarche was based on self-reported presence/absence of the first menstrual cycle. Pregnancy status was determined during medical visits based on the date of last menses, with urinary pregnancy tests administered by the physician when pregnancy was suspected. Pregnancies were cross-validated against subsequent annual demographic and census interviews, allowing detection of pregnancies that occurred between medical visits and pregnancies that went undetected during previous physician examinations [@Blackwell2015c]. Menopause was determined based on reported absence of a menstrual cycle over the past 6 months.

To obtain a representative sample from the United States, we used publicly available cross-sectional NHANES data collected by the Centers for Disease Control (https://www.cdc.gov/nchs/nhanes/index.htm) between 2003 and 2016. Total and differential leukocyte counts were measured using the Coulter method. Females who were under the age of 8 (for which reproductive data were redacted) or who specifically reported absence of menarche were binned as premenarchal. Females who were not currently pregnant, had not given birth within the preceding 12 months, and who reported a regular menstrual cycle either at time of exam or within the preceding two months were binned as regularly cycling. Females who self-reported being pregnant and/or had a positive urine test at the time of exam were categorized as currently pregnant. Women who were not currently pregnant but had given birth within the past 12 months were considered postpartum. Finally, females who reported absence of regular menstruation due to menopause were binned as postmenopausal.

### Sample Selection
Final THLHP and NHANES data sets were limited to males and females ages 2 to 84 with recorded white blood cell differential and body mass index (BMI). NLR was calculated by dividing neutrophil count by total lymphocyte count. We excluded `r early_meno$n` females from the NHANES dataset who reported reaching menopause before age 40 and `r late_meno$n` individual who reported regular menstruation after the age of 70 (effectively removing outliers associated with very early versus delayed menopause and/or errors in the data). We also removed extreme values for immune cell counts and NLR by limiting our sample to values which fell between population-pooled 1% and 99% percentiles. This exclusion criterion resulted in `r sum(removed_data$removed)` data points (`r str_c(percent = format(round(sum(removed_data$removed)/sum(removed_data$original)*100, digits = 2), nsmall = 2), "%")`) being removed out of the initial pooled dataset consisting of `r sum(removed_data$original)` unique values. Given the ubiquity of chronic parasitic infections among the Tsimane, we did not exclude individuals with known infection or illness. In both populations, BMI varies considerably by age and sex (Figure S1), so we z-scored all BMI values using sex and age-specific mean values. No exclusions were based on medical diagnoses. Finally, we used the *matchit* package [@ho_etal11] to create matched THLHP and NHANES samples stratified by age, sex, and female reproductive phase (specifying nearest neighbor matching on propensity score) (Figure S2, Table S1). 

### Statistical Analyses
All models were executed in R 4.3.2 (https://cran.r-project.org) using the *mgcv* package [@wood11]. We employed generalized additive models (GAMs) to estimate the non-linear population-specific effects of age by sex and female reproductive phase (premenarchal, regularly cycling, pregnant, postpartum, postmenopausal) on total leukocyte, neutrophil, total lymphocyte, eosinophil, and monocyte count and neutrophil-to-lymphocyte ratio (NLR). All models also accounted for the fixed effects of z-scored BMI and the smoothed effects of parity, stratified by female reproductive phase (males and premenarchal females were all assigned parity values of zero). Due to repeat sampling of individuals within the THLHP dataset (Table S2), all THLHP-specific models accounted for the group-level random effects of participant identification number. 

\newpage

## Figures
<!-- Provide all images at final size. While figures may be sized conservatively to save page space, PNAS reserves the right to make the final decision on figure size in published articles and authors may be asked to shorten manuscripts that exceed the stated length requirement. -->
<!-- Small: approximately 9 cm x 6 cm -->
<!-- Medium: approximately 11 cm x 11 cm -->
<!-- Large: approximately 18 cm x 22 cm -->
<!-- Ensure that all numbers, letters, and symbols are no smaller than 6 points (2 mm) and no larger than 12 points (6 mm) after reduction. Keep text sizing consistent within each graphic. -->
<!-- Preassemble all composite figures.  -->
<!-- Submit images in these file formats: TIFF, EPS, PDF, or PPT. -->

<!-- Include figure legends immediately after referencing the figure in the manuscript. Ensure that figure legends adhere to the following guidelines:  -->

<!-- For figures with multiple panels, the first sentence of the legend should be a brief overview of the entire figure. Explicitly reference and describe each panel at least once in the figure legend. -->
<!-- Include clearly labeled error bars in all graphs and describe them in the figure legend. -->
<!-- State whether a number that follows the ± sign is a standard error (SEM) or a standard deviation (SD). -->
<!-- When applicable, provide the P value, magnification, or scale bar information.  -->
<!-- Indicate the number of independent data points (N) represented in a graph in the legend. -->
<!-- Ensure that numerical axes on all graphs go to 0, except for log axes. -->

<!-- Combination halftones: Common examples are color or grayscale figures containing halftone and line art -->
<!-- elements. The preferred resolution for this type of image is between 600 and 900 ppi at publication size.  -->

![Fig 1. Non-linear effects of age by sex and female reproductive phase on (A) leukocyte differential among the Tsimane, (B) leukocyte differential in the USA, (C) NLR among the Tsimane, and (D) NLR in the USA (Table 1). Estimated values are standardized by parity (set to 0 live births for premenarchal females and 3 live births for regularly cycling, pregnant, postpartum, and postmenopausal females) and z-scored BMI (set to 0). Solid lines correspond to estimated mean value, shaded regions correspond to 95% credible intervals. Please note that each variable is plotted on its own scale.](../output/figure_1.png){width=100%}

\newpage

![Fig 2. Non-linear effects of parity by female reproductive phase on (A) leukocyte differential among the Tsimane, (B) leukocyte differential in the USA, (C) NLR among the Tsimane, and (D) NLR in the USA (Table 1). Estimated values are standardized by age (set to 24 years for regularly cycling, pregnant, and postpartum females and set to 65 years for postmenopausal females) and z-scored BMI (set to 0). Solid lines correspond to estimated mean value, shaded regions correspond to 95% credible intervals. Please note that each variable is plotted on its own scale.](../output/figure_2.png){width=80%}

\newpage

## Tables
<!-- Tables -->
<!-- Ensure that the table is in an editable Word, RTF, or LaTeX format. -->
<!-- Include a brief title (above) and footnotes (below) the table. -->
<!-- Avoid multipart tables (Table 1A, Table 1B). -->

```{r, echo=F}
knitr::kable(table_1, 
             caption = "Table 1. Description of immune markers used in this study.")
```

\newpage

```{r, echo=F}
knitr::kable(table_2 %>% 
               mutate(`Reproductive Phase` = case_when(`Reproductive Phase` == "Pre-Menst" ~ "Premenarchal",
                                           `Reproductive Phase` == "Menopause" ~ "Postmenopausal",
                                           TRUE ~ `Reproductive Phase`),
                      `Reproductive Phase` = ordered(`Reproductive Phase`,
                            levels = c("Premenarchal","Cycling","Pregnant",
                                       "Postpartum","Postmenopausal"))) %>%
               select(-Group) %>%  
               arrange(`Reproductive Phase`, Population,Measure) %>% 
               select(-`Age Span`, -`% Age Span`), 
             caption = "Table 2. Ages at which the 95% credible intervals for estimated values between males and females do not overlap, separated by population, measure, and female reproductive phase. Estimated values presented in this table are standardized by parity (set to 0 live births for premenarchal females and 3 live births for regularly cycling, pregnant, postpartum, and postmenopausal females) and z-scored BMI (set to 0).")
```

\newpage

## Data availability
R code for the statistical analyses is published on a public GitHub repository (https://github.com/carmenhove/sex_differences). Individual-level data are stored in the THLHP Data Repository and are available through restricted access for ethical reasons. THLHP's highest priority is the safeguarding of human subjects and minimization of risk to study participants. The THLHP adheres to the "CARE Principles for Indigenous Data Governance" (Collective Benefit, Authority to Control, Responsibility, and Ethics), which assure that the Tsimane (i) have sovereignty over how data are shared, (ii) are the primary gatekeepers determining ethical use, (iii) are actively engaged in the data generation, and (iv) derive benefit from data generated and shared for use whenever possible. The THLHP is also committed to the "FAIR Guiding Principles for scientific data management and stewardship" (Findable, Accessible, Interoperable, Reusable). Requests for individual-level data should take the form of an application that details the exact uses of the data and the research questions to be addressed, procedures that will be used for data security and individual privacy, potential benefits to the study communities, and procedures for assessing and minimizing stigmatizing interpretations of the research results (see the following webpage for links to the data sharing policy and data request forms: https://tsimane.anth.ucsb.edu/data.html). Requests for individual-level data will require institutional IRB approval (even if exempt) and will be reviewed by an Advisory Council composed of Tsimane community leaders, community members, Bolivian scientists, and the THLHP leadership. The study authors and the THLHP leadership are committed to open science and are available to assist interested investigators in preparing data access requests.

## Acknowledgements
We offer our profound thanks to all participants who participated in this study, the Tsimane Gran Consejo, and the THLHP mobile team and staff.

## Funding
The THLHP was supported by the National Institute of Health/National Institute on Aging (R01AG054442, R01AG024119, R56AG024119, P01AG022500) and National Science Foundation (BCS0422690). J.S. acknowledges Institute for Advanced Study in Toulouse funding from the French National Research Agency (ANR). 

\newpage

## References
<div id="refs"></div>
