---
title: "notes"
output: html_document
date: "2024-04-04"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r manuscript setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
pacman::p_load(devtools, tidyverse, openxlsx, ggpubr, 
               brms, calecopal, citr, knitr, wordcountaddin)

load('pnas.RData')
```

<!-- PNAS is format-neutral at initial submission, which means that manuscripts do not need to be formatted according to specific journal guidelines to be considered for review. We do, however, require the following information in order to evaluate your manuscript: -->

<!-- a manuscript file (in any format) including the following: -->
<!-- title page (title, author list, classification, keywords) -->
<!-- abstract -->
<!-- significance statement -->
<!-- main text -->
<!-- references -->
<!-- figures or tables with appropriate legends (may be uploaded separately) -->
<!-- SI files (may be uploaded separately) -->
<!-- contact and competing interest information for all authors -->
<!-- data sharing plans (for all data, documentation, and code used in analysis) -->
<!-- funding information and whether an open access license has been selected -->
<!-- list of appropriate Editorial Board members, NAS members, and qualified reviewers (minimum of three each) who are experts in the paper's scientific area. (A brief justification for suggested reviewers is welcome, particularly for interdisciplinary papers) -->

<!-- Title page -->
<!-- Abstract -->
<!-- Explain to the general reader the major contributions of the article -->
<!-- Include no more than 250 words -->
<!-- Cite all references in the abstract in full within the abstract itself AND in the text -->
<!-- Significance statement (Direct and Contributed Submissions only) -->
<!-- Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty -->
<!-- Include no more than 120 words -->
<!-- Main text -->
<!-- Introduction -->
<!-- Results -->
<!-- Discussion -->
<!-- Materials and methods (describe procedures in sufficient detail so that the work can be repeated) -->
<!-- Acknowledgments and funding sources -->
<!-- Spell out all abbreviations -->
<!-- Use FundRef to identify the standard name for any funders -->
<!-- Do not include dedications -->
<!-- References -->
<!-- Figure legends -->

## Title Page

### Title: Female reproductive state and ecological conditions impact the magnitude of sex differences in immune status across the lifespan

### Authors
<!-- Author affiliations -->
<!-- Include, in this order, department/laboratory/section/division, institution, city, state with ZIP code (for US institutions) or country with postal code (for non-US institutions). -->
<!-- Use superscripts to match authors with institutions. Multiple affiliations are allowed. Affiliations where the work was done should be listed. If an author's affiliation has changed since the work was done, the current affiliation should also be listed and noted as such. -->
<!-- PNAS strongly encourages authors to supply an ORCID identifier for each author. Do not include ORCIDs in the manuscript file; individual authors must link their ORCID account to their PNAS account at www.pnascentral.org. For proper authentication, authors must provide their ORCID at submission and are not permitted to add ORCIDs on proofs. -->

Carmen Hove\*^1^, Michael D Gurven^1^, Benjamin C Trumble^2^, Jonathan Stieglitz^3^, Daniel Eid Rodriguez^4^, Ivan Maldonado Suarez^5^, Hillard Kaplan^6^, Amy Boddy^1^, Aaron D Blackwell\*^7^ 

^1^ Department of Anthropology, University of California Santa Barbara, Santa Barbara, California, 93106 \
^2^ Center for Evolution and Medicine, School of Human Evolution and Social Change, Institute of Human Origins, Arizona State University, Tempe, Arizona, 85287 \
^3^ Institute for Advanced Study in Toulouse, Toulouse School of Economics, Toulouse, France, 31015 \
^4^ Department of Medicine, Universidad de San Simón, Cochabamba, Bolivia \
^5^ Tsimane Health and Life History Project, San Borja, Bolivia \
^6^ Economic Science Institute, Chapman University, Orange, California, 92866 \
^7^ Department of Anthropology, Washington State University, Pullman, Washington, 99164 \

Corresponding Authors\* 

Carmen Hove \
Email: carmenhove@ucsb.edu \
Phone: (360) 621-3508 \

Aaron Blackwell \
Email: aaron.blackwell@wsu.edu \
Phone: (805) 253-2796

<!-- Carmen Hove* \ -->
<!-- ORCID: https://orcid.org/0000-0002-8626-321X \ -->

<!-- Michael Gurven \ -->
<!-- ORCID: https://orcid.org/0000-0002-5661-527X \ -->
<!-- Email: gurven@anth.ucsb.edu \ -->

<!-- Ben Trumble \ -->
<!-- ORCID: https://orcid.org/0000-0003-3201-0628 \ -->
<!-- Email: trumble@asu.edu \ -->

<!-- Jonathan Stieglitz \ -->
<!-- ORCID: https://orcid.org/0000-0001-5985-9643 \ -->
<!-- Email: jonathan.stieglitz@gmail.com \ -->

<!-- Daniel Eid Rodriguez -->
<!-- ORCID: https://orcid.org/0000-0001-8145-0967 -->

<!-- Ivan Maldonado Suarez -->
<!-- ORCID: https://orcid.org/0009-0003-7345-6519 -->

<!-- Hillard Kaplan \ -->
<!-- ORCID: https://orcid.org/0000-0002-7398-7358 \ -->
<!-- Email: hkaplan@chapman.edu \ -->

<!-- Amy Boddy \ -->
<!-- ORCID: https://orcid.org/0000-0002-7990-2415 \ -->
<!-- Email: boddy@anth.ucsb.edu \ -->

<!-- Aaron Blackwell* \ -->
<!-- ORCID: https://orcid.org/0000-0002-5871-9865 \ -->
<!-- Affiliation: Washington State University -->

### Classification
<!-- Select both a major and minor classification category. Dual classifications are permitted between major categories. -->

Major: Biological Sciences \
Minor: Immunology and Inflammation

### Keywords
<!-- Include at least three keywords at submission. -->

\newpage

### Abstract
<!-- Explain to the general reader the major contributions of the article. Include no more than 250 words. -->

In industrialized societies, females benefit from lower infectious disease burden and lower risk of non-reproductive cancers than males but suffer disproportionately from autoimmune disease. Considering the role that sex hormones play in shaping sex differences in immune function and the sensitivity of female ovarian function to energetic conditions, evolutionarily novel conditions common to industrialized populations (e.g., energetic excess, reduced parity) may exacerbate evolved sex differences in immunity. To test this hypothesis, we estimate and compare the age-specific effects of sex and female reproductive state on leukocyte differential and neutrophil-to-lymphocyte ratio (NLR) among the Tsimane (a non-industrialized population) and a representative sample from the USA. We find that the magnitude of sexual dimorphism in immune measures across the lifespan is substantially lower among the Tsimane when compared to the USA. These population-level differences are largely driven by comparatively large effects of pregnancy and menopause on immune status among females in the USA. Low-parity pregnant females in the USA exhibited substantially higher neutrophil count and NLR than their age-matched male peers (and regularly cycling and postpartum females), while postmenopausal females in the USA exhibited substantially lower neutrophil, eosinophil, and monocyte counts and NLR relatively to their age-matched male counterparts. These findings suggest that multiple pregnancies and the transition to menopause exert smaller effects on female immune status in non-industrialized contexts, resulting in attenuated sex differences in immune status across the lifespan. 

### Significance Statement
<!-- Explain the significance of the research at a level understandable to an undergraduate-educated scientist outside their field of specialty. Include no more than 120 words. -->

Current research indicates that women generally mount greater immune defenses against infectious agents but suffer increased risk of autoimmune disease. Given that most research on sex differences in immune function comes from industrialized urban populations and often excludes women in different reproductive phases (e.g., pregnancy, menopause), it is unclear how magnitude of sex differences in immune status is influenced by socio-ecological conditions impacting sex hormone production (e.g., energetic status, parity). Here we show that the magnitude of sexual dimorphism in immune measures across the lifespan is substantially lower among the Tsimane (a natural-fertility non-industrialized population) when compared to the USA and that this pattern is primarily driven by population-specific effects of pregnancy and menopause.

\newpage

## Main Text

### Introduction

In post-industrial human populations, women benefit from reduced infectious disease burden and lower risk of non-reproductive cancers relative to men but make up approximately 80% of autoimmune disease diagnoses [@Whitacre2001a] and suffer disproportionately from allergy and atopy [@Laffont2017]. These sex biases in disease risk are reflected in sex differences across immune biomarkers [@Klein2016]. Women generally possess more neutrophils [@bain_england75; @casimir_etal10; @chen_etal16a; @li_etal20], granulocytes that orchestrate first-line defenses against pathogens, prime antigen-specific immunity, and induce potentially deleterious inflammatory cascades [@chen_etal14; @mayadas_etal14; @polak_etal19]. Women also have higher B cell and CD4^+^ T cell counts [@abdullah_etal12; @uppal_etal03]. Conversely, total lymphocyte count is often higher in males [@abdullah_etal12], reflecting higher numbers of regulatory T [@afshan_etal12] and natural killer cells [@abdullah_etal12]. Consequently, women frequently possess a comparatively higher neutrophil-to-lymphocyte ratio (NLR). NLR is a robust indicator of immune homeostasis and systemic inflammation [@kourilovitch_galarza-maldonado23] strongly associated with the presence and severity of numerous autoimmune disorders that disproportionately afflict women (e.g., rheumatoid arthritis) [@zhou_etal23; @damico_etal19; @hasselbalch_etal18; @mihai_etal22; @wang_etal20]. While there is ongoing debate regarding the ultimate selective pressures responsible for such sex differences in immune function [@McKean2005; @metcalf_graham18a; @fink_klein18; @mitchell_etal22; @Natri2019], there is robust evidence that sex-specific hormone production [@kocar_etal01; @gayen_etal16; @Taneja2018; @furman_etal14] is an evolutionarily conserved proximate mechanism by which females mount stronger cellular and humoral responses to pathogens [@Fish2008b; @Takahashi2020], vaccination [@klein_etal10; @Mitchell1999; @Voigt2019], and auto-antigens.  

<!-- Proposed selective pressures responsible for producing such sex differences in immune function include mate choice and the corresponding development and maintenance of "costly" secondary male sex characteristics [@McKean2005], sex-specific reproductive schedules [@metcalf_graham18a], the need for increased transfer of antibodies to offspring during and after pregnancy [@fink_klein18], vertical transmission of pathogens from mother to offspring during pregnancy [@mitchell_etal22], and "the presence of an invasive placenta and an immunologically challenging pregnancy" (Pregnancy Compensation Hypothesis) [@Natri2019]. -->

<!-- incomplete X-inactivation of immune-related genes during female development [@gayen_etal16] and  -->

Because female hormone production vacillates as function of current reproductive state (premenarchal, regularly menstruating, pregnant, postpartum, and postmenopausal), the direction and magnitude of sex biases in immune function should also vary by female reproductive status. The impact of such *within-sex* variability on overall sex differences in immune status is not well understood, however. Likewise, the effect of cumulative reproductive effort (e.g., total number of pregnancies) on female immune function and corresponding sexual dimorphism in immune status is under-explored. The Pregnancy Compensation Hypothesis (PCH) proposes that reduced parity amplifies sex differences in immune function by creating a mismatch between 'compensatory' immunological responses (mediated by hormones and required for immune competence during pregnancy) and less time spent gestating over the lifespan [@Natri2019]. To date, we know of no studies that have tested this prediction. Furthermore, most studies on sex differences in immune function come from post-industrial or rapidly industrializing societies, where evolutionarily novel conditions (e.g., microbial deprivation, low energetic throughput) may exacerbate evolved sex differences in immunity by limiting opportunities for immunological calibration during development and increasing exposure and/or sensitivity to sex hormones. In this study, we investigate these questions by estimating and comparing the age-dependent effects of sex, stratified by female reproductive state, on immune cell counts and neutrophil-to-lymphocyte ratio within a heavily industrialized society (the USA) and a natural-fertility subsistence population (the Tsimane). Among postmenarchal females, we also estimate and compare the population-specific effects of parity on immune measures, stratified by female reproductive state, to test the PCH's prediction that parity is negatively associated with overall magnitude of sex bias in immune status [@Natri2019]. 

<!-- Female reproduction: which phase, at what age, how many times, and in what context.  -->

#### Female reproduction requires substantial hormonal and immunological shifts
 
While general aging impacts hormone production [@handelsman_etal16; @decaroli_rochira17] and immune senescence [@marquez_etal20] in both sexes, female reproduction requires additional hormonal changes and presents unique immunological challenges. During pregnancy, production of estradiol, estrone, testosterone, and progesterone are significantly elevated above baseline [@oleary_etal91; @schock_etal16]. These hormonal shifts support fetal development [@magon_kumar12] and play a critical role in inducing fetal tolerance, a phenomenon in which the maternal immune system must tolerate fetal antigens while maintaining essential immune defenses [@schumacher_etal14a; @soma-pillay_etal16]. Following delivery, ovarian function shifts again: estrogen and progesterone production are suppressed by the antagonistic effects of prolactin, a hormone which remains elevated in the presence of regular lactation [@neville_neifert83]. There is some evidence that pregnancy produces lasting alterations in the production of certain hormones, including reduced prolactin secretion [@musey_etal87a] and elevated estriol [@musey_etal87], but how these long-term alterations impact immune function is not well understood. Finally, menopause marks the end of the female reproductive lifespan, with the ovaries switching to low production of estrogens and progesterone and continued production of androgens (**citations**) [@burger02]. Although males do not experience an exact physiological equivalent to menopause, advancing age in men is also marked by significant declines in sex hormone production [@decaroli_rochira17] and overall fertility [@khandwala_etal18; @pino_etal20].  

<!-- Even in the absence of pregnancy, female sex hormone levels fluctuate across the ovarian cycle in preparation for implantation, although not to the extent that they do during gestation (**citations**).  -->

The dynamic nature of ovarian function over the life course suggests that hormone-mediated sex biases in immune function and disease risk should be most pronounced between puberty and menopause and may be accentuated by current (and possibly prior) pregnancy. In line with this prediction, autoimmune diseases that most disproportionately affect women (e.g., primary Sjögren's syndrome, systemic lupus erythematosus, Hashimoto thyroiditis) are most commonly diagnosed among women between 20 and 50 years of age [@garcia-carrasco_etal02; @ohta_etal13; @mincer_jialal24; @moinzadeh_etal20; @aarli99] and often emerge or worsen after pregnancy [@piccinni_etal16]. While these patterns in overall disease risk provide some insight, regular exclusion of pregnant women, inconsistent binning of participants into categorical age groups, and failure to consider the effects of parity preclude a clear picture of how age, current female reproductive status, and female reproductive history combine to impact sex differences in immune status.

<!-- [@bain_england75; @hu_etal20a; @li_etal20; @pellegrino_etal23; @thammawong_etal]  -->

<!-- Male and female hormone profiles begin to differentiate in utero [@carr_etal83] and continue to develop during infancy and childhood [@karaoglan_nacarkahya21; @moreira-filho_etal18], with puberty marking the transition to reproductive maturity and the establishment of adult-level hormone production in both sexes.  -->

<!-- To date, however, regular exclusion of pregnant women, inconsistent binning of participants into categorical age groups, and failure to consider the effects of parity [@bain_england75; @hu_etal20a; @li_etal20; @pellegrino_etal23; @thammawong_etal] preclude a clear picture of how age, current female reproductive status, and female reproductive history combine to impact sex differences in immune status.  -->


<!-- Even in the absence of pregnancy, female sex hormone levels fluctuate across the ovarian cycle in preparation for implantation, although not to the extent that they do during gestation (**citations**).  -->

<!-- [@bain_england75; @hu_etal20a; @li_etal20; @pellegrino_etal23; @thammawong_etal]  -->

<!-- Male and female hormone profiles begin to differentiate in utero [@carr_etal83] and continue to develop during infancy and childhood [@karaoglan_nacarkahya21; @moreira-filho_etal18], with puberty marking the transition to reproductive maturity and the establishment of adult-level hormone production in both sexes.  -->

<!-- To date, however, regular exclusion of pregnant women, inconsistent binning of participants into categorical age groups, and failure to consider the effects of parity [@bain_england75; @hu_etal20a; @li_etal20; @pellegrino_etal23; @thammawong_etal] preclude a clear picture of how age, current female reproductive status, and female reproductive history combine to impact sex differences in immune status.  -->

#### Hormone production and immunological development are sensitive to ecological inputs

Socio-ecological conditions that increase overall energy balance (e.g., caloric excess, increased sedentism) may magnify hormone-mediated sex differences in immune function via disproportionately large effects on female ovarian function. Because female reproduction across mammalian species requires significantly greater energetic investment than male reproduction, female ovarian function is particularly responsive to energy balance. While reduced energetic availability suppresses baseline ovarian function, resulting in lower progesterone production across the ovarian cycle [@Jasienska1998a; @nunez-delamora_etal07; @vitzthum_etal04a], positive energy balance and reduced metabolic load are linked to earlier age at menarche [@thomas_etal01], elevated adult levels of estradiol [@emaus_etal08] and progesterone [@nunez-delamora_etal07], higher progesterone levels during the peri-ovulatory and peri-implantation period [@vitzthum_etal04a], and shorter duration of lactational amenorrhea among breastfeeding women [@valeggia_ellison09]. While moderate energetic availability is associated with enhanced fecundity [@sadhir_pontzer23] and greater immune competence [@childs_etal19], caloric excess contributes to chronic inflammation and greater risk of hypersensitivity to a broad range of antigens [@macsali_etal09]. Furthermore, exposure to *exogenous* female sex hormones via regular use of oral contraceptives, an evolutionarily novel feature of industrialized societies, has also been linked to immunological hypersensitivity to a variety of antigens - including steroid hormones themselves [@untersmayr_etal17]. 

<!-- While testosterone production among adult males responds to energy expenditure [@trumble_etal23; @nindl_etal07; @spratt_etal06; @muehlenbein_etal10a] and is therefore generally elevated in industrialized populations [@bribiescas96], interpopulation variation in male testosterone levels is not as large as variation in ovarian hormone production among females [@Jasienska1998a; @nunez-delamora_etal07; @vitzthum_etal04a].  -->

Lack of exposure to certain pathogens during development may also increase sexual dimorphism in immune function, not just by increasing the amount of energy allocated to sex-specific hormone production, but by reducing the number of opportunities for immunological calibration. In the United States, for example, reductions in infectious disease burden [@armstrong_etal99; @roush07] have increased life expectancy [@cutler_miller05] but are linked to the rise of chronic inflammatory disorders characterized by hypersensitivity to non-pathogenic antigens (e.g., atopy, autoimmune disease) [@bloomfield_etal16]. One explanation for this relationship is that exposure to pathogens during development primes the immune system to differentiate between pathogenic and non-pathogen antigens and, in certain cases, induce immunological tolerance. People living in a non-industrialized context characterized by elevated pathogen load exhibit higher baseline levels of most immune markers [@Blackwell2011c; @Blackwell2016] but seemingly low incidence of allergy or autoimmune disease [@gurven_etal07; @Gurven2008a; @kaplan_etal17; @wu_etal23]. Lack of calibrating opportunities via pathogen exposure may have particular relevance during pregnancy, when the maternal immune system must induce tolerance of fetal antigens. Current evidence indicates that non-industrialized populations experiencing high pathogen exposure exhibit less immunological activation (e.g., lower peak in neutrophil count and C-reactive protein) during pregnancy [@Hove2020], which may reflect both lower hormone production and reduced immunological sensitivity to fetal antigens during gestation. 

Taken together, these patterns suggest that evolutionarily novel environmental conditions common to industrialized populations (e.g., low energetic throughput, microbial deprivation) may exacerbate evolved sex differences in immunity by reducing opportunities for immunological calibration during development and increasing exposure and/or sensitivity to sex hormones - especially among females. At time of writing, however, these ideas remain largely unexplored and therefore current definitions of "normal" sexual dimorphism in immune status rest primarily on studies conducted within industrialized/industrializing societies (e.g., China, the USA, India).

<!-- Finally, the proposed dampening effects of multiple pregnancies on female immune function, as posited by the Pregnancy Compensation Hypothesis [@Natri2019], may also vary as a function of energetic constraints. There is some evidence that women in industrialized societies exhibit higher progesterone levels during the peri-ovulatory and peri-implantation period [@vitzthum_etal04a] and greater increase in neutrophil count and C-reactive protein during pregnancy [@Hove2020] when compared to women in non-industrialized populations. Furthermore, women in industrialized countries experience shorter duration of lactational amenorrhea postpartum, even while breastfeeding, compared to those living in more resource-restricted environments [@valeggia_ellison09]. The effects of each additional pregnancy on hormone levels and downstream immune status   -->

<!-- Taken together, these patterns suggest that relatively unfettered sex hormone production and lack of calibrating opportunities (e.g., fewer encounters with helminths) may increase immunological *responsiveness* to potential allergens and auto/semi-auto antigens and therefore contribute to the high incidence of autoimmune disease observed among women in industrialized populations. -->

#### Objectives and predictions

<!-- Specifically, we use generalized additive models (GAMs) to determine the population-specific smooth effects of age, stratified by sex and female reproductive state (premenarchal, regularly cycling, pregnant, 0-12 months postpartum, and postmenopausal), on leukocyte differential and neutrophil-to-lymphocyte ratio (Table 1). Among all postmenarchal females, we also estimate the population-specific effects of parity on immune measures, stratified by population and female reproductive state, to test the PCH's prediction that parity is negatively associated with overall magnitude of sex bias in immune status [@Natri2019]. -->

Here, we utilize data from the Tsimane (a natural-fertility subsistence population inhabiting the Amazonian River basin) and a representative sample from the United States to estimate the age-dependent effects of sex, stratified by female reproductive phase, on white blood cell counts and neutrophil-to-lymphocyte ratio (Table 1). Among postmenarchal females, we also estimate the population-specific effects of parity on these immune markers, stratified by female reproductive state. We use white blood cell differential and neutrophil-to-lymphocyte ratio as our markers of immune status due to the breadth of immunological processes these biomarkers reflect [@heim_etal18; @ishihara_etal06] and their diagnostic and prognostic utility across numerous health outcomes [@al-gwaiz_babay07; @kourilovitch_galarza-maldonado23; @lowsby_etal15].  

Using this approach, we evaluate the following predictions:

* Within both populations, sex biases in immune status will be most pronounced between postmenarchal and premenopausal females and their age-matched male peers, with pregnancy corresponding to further increased magnitude of sexual dimorphism. Given the reversal of pregnancy-induced hormonal patterns, suppression of ovulation, and reduced exposure to fetal antigens that follows delivery, postpartum females will exhibit less divergent immune profiles from their male counterparts than their regularly cycling or pregnant female peers. Likewise, sex differences in immune status will be attenuated among postmenopausal women and their male peers, due to the drop-off in ovarian hormone production during menopause. 
* Within both populations, there will be a negative relationship between parity and degree of sex bias in immune function among parous postmenarchal females [@Natri2019].
* Across all ages and female reproductive states, sex biases in immune measures will be attenuated among the Tsimane compared to the USA. 

<!-- We build on this prediction by proposing that, given current evidence on population differences in hormonal changes preceding implantation [@vitzthum_etal04a] and immune responses during gestation [@Hove2020], the dampening effect of consecutive pregnancies on sex differences in immune status may be smaller in the USA.  -->

### Results
<!-- We compare immune profiles of Tsimane and US women and men across four stages (premenarche, cycling, pregnancy, postmenopause). -->

#### Female reproductive status has age-dependent and population-specific effects on the direction and magnitude of sex differences in immune markers
<!-- Make tables for age, reproductive status, and parity effects -->

Among males in both populations, age is associated with strong non-linear and population-specific effects on all immune measures - especially within the first two decades of life (Figure 1, Table S4, Table S5). Among females, the effects of age on immune measures are mediated by current reproductive state and differ based on population and immune measure (Figure 1, Table S4, Table S5). 

In the USA, premenarchal females have lower eosinophil counts than males between the ages of `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Max"]`, lower monocyte counts between the ages of `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Max"]`, and slightly higher total lymphocyte counts from ages `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Max"]` and again at age `r get.values(table_2, "NHANES","LYM", "Pre-Menst", 2)[,"Min"]`. We find no differences in neutrophil count, total white blood cell count, and NLR  between premenarchal females and their male peers within the USA, regardless of age (Figure 1A, Figure 1C, Table 2). Similar to premenarchal females, regularly cycling females in the USA have lower eosinophil counts than males at all ages and lower monocyte counts between ages `r get.values(table_2, "NHANES","MON", "Cycling")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Cycling")[,"Max"]` (Figure 1A). As shown in Table 2, regularly cycling females in the USA have higher neutrophil and total white blood cell counts than males at younger and later ages, but most ages are characterized by an *absence* of robust sex biases in these immune measures. Likewise, there is a female bias in NLR between regularly cycling females in the USA and their male counteparts, but only from ages `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Min"]` to `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Max"]`. In contrast, pregnant females in the USA have higher neutrophil count, total white blood cell count, and NLR than their male peers for the majority of ages represented, with highly non-linear effects of age producing maximum sex differences at `r get.abs.values(table_s3, "NHANES","NEU","Pregnant")[,"Age"]`, `r get.abs.values(table_s3, "NHANES","WBC","Pregnant")[,"Age"]`, and `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Age"]` years of age, respectively (Figure 1A, Table 2, Table 3). Conversely, pregnant females in the USA have lower total lymphocyte count than males from ages `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Min"]` to `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Max"]` and lower eosinophil counts between ages `r get.values(table_2, "NHANES","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Max"]`. Postpartum females in the USA have lower monocyte counts than males from ages `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Max"]` but otherwise exhibit no differences from their male counterparts. Finally, depending on age, postmenopausal females in the USA have *lower* neutrophil, eosinophil, and monocyte counts than males and *higher* total lymphocyte count (Figure 1A, Table 2). 

Among the Tsimane, premenarchal females have slightly *lower* total lymphocyte counts, but only from ages `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Max"]`. Likewise, premenarchal females possess slightly *higher* monocyte counts than males, but only from ages `r get.values(table_2, "THLHP","MON", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "THLHP","MON", "Pre-Menst")[,"Max"]`. Among individuals at younger ages, regularly cycling Tsimane women have lower neutrophil, total lymphocyte, eosinophil, and total white blood cell counts and higher NLR than males - but most ages are characterized by an *absence* of robust sex biases in these immune measures (Figure 1B, Table 2). Pregnant Tsimane females have lower total lymphocyte count than males from ages `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Min"]` to `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Max"]`, lower eosinophil counts between ages `r get.values(table_2, "THLHP","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Max"]`, and lower total white blood cell count from ages `r get.values(table_2, "THLHP","WBC", "Pregnant")[,"Min"]` to `r get.values(table_2, "THLHP","WBC", "Pregnant")[,"Max"]`. Postpartum females in the Tsimane have *higher* monocyte counts than males from ages `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Min"]` and `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Max"]` but otherwise exhibit no differences from their male counterparts. Finally, depending on age, postmenopausal females in the Tsimane have *lower* NLR and eosinophil counts than males (Figure 1, Table 2).

#### Higher parity corresponds with reduced sex bias in neutrophil count and NLR, but only among pregnant women in the USA
We do not find any robust effects of parity on immune markers among regularly cycling, pregnant, or postpartum Tsimane females. Likewise, we do not find strong effects of parity on immune measures among regularly cycling or postpartum females in the USA (Figure 2, Table S4, Table S5). We do, however, find strong negative effects of parity on neutrophil count `r get.statistics(model_summaries, "USA","NEU","s(Parity):Pregnant")`, monocyte count `r get.statistics(model_summaries, "USA","MON","s(Parity):Pregnant")`, and NLR `r get.statistics(model_summaries, "USA","NLR","s(Parity):Pregnant")` among pregnant females in the USA (Figure 2, Table S5). Consequently, high-parity pregnant women in the USA exhibit immune profiles that are more similar to age-matched men compared to primiparous women. Among postmenopausal females in both populations, we find a non-linear but increasingly positive effect of parity on NLR (Figure 2, Table S4, Table S5). Among postmenopausal Tsimane females, we observe a negative effect of parity on eosinophil count `r get.statistics(model_summaries, "Tsimane","EOS","s(Parity):Postmenopausal")`. 

#### The direction and magnitude of sexual dimorphism in immune status varies between the USA and the Tsimane

The results described above create age-dependent differences in both the direction and magnitude of sexual dimorphism in immune cell counts and NLR between the Tsimane and the USA. Among premenarchal females and their male peers, there is a female bias in total lymphocyte count and male bias in eosinophil and monocyte counts in the USA but only a short-lived *male* bias in total lymphocyte count among the Tsimane. Among regularly cycling females and their male peers, there is a female bias in neutrophil count, NLR, and total white blood cell count and sustained male bias in eosinophil and monocyte counts in the USA versus a *male* bias in neutrophil, total lymphocyte, and white blood cell count among the Tsimane. Among pregnant females and their male counterparts, we find a female bias in NLR and a male bias in total lymphocyte and eosinophil counts in both populations. In the USA, however, female bias in NLR is further elevated by a corresponding female bias in neutrophil count that we do not find among the Tsimane. Among postpartum females and their male peers, we find an age-dependent male bias in monocyte count in the USA and an age-dependent female bias in monocyte count among the Tsimane. Among postmenopausal females and their male peers,there is a female bias in total lymphocyte count and male bias in neutrophil count, eosinophil count, monocyte count, and NLR in the USA. Among the Tsimane, the only sex bias we observe among postmenopausal women and age-matched men is a slight male bias in NLR among elderly individuals 

 <!-- we find more robust and longer-lasting sexual dimorphism among postmenopausal females and their male peers in the USA compared to the Tsimane -->
 
<!-- Controlling for parity, the greatest absolute sex difference in NLR between pregnant women in the USA and their male counterparts occurs at age `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Age"]`, with a female ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Male Value"]` - an approximately `r 3.15-2.01` difference. The greatest absolute sex difference in NLR between pregnant Tsimane women and their male counterparts also occurs at age `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Age"]`, but with a female ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Male Value"]` - an approximately `r 2.43-1.79` difference.  -->

<!-- In the USA, we observe a robust female bias in neutrophil count between pregnant females and their age-matched male counterparts (moderated by age and parity). Conversely, we find no robust sex bias in neutrophil count between pregnant Tsimane females and their male peers, regardless of age or parity. Similarly, there is an age-dependent male bias in neutrophil count between postmenopausal females and males in the USA, but no evidence of sex differences among the postmenopausal females and males among the Tsimane.  -->

<!-- In the USA, we find an age-dependent *female* bias in total lymphocyte count between premenarchal females and their male counterparts. Among the Tsimane, we find an age-dependent *male* bias in total lymphocyte count between premenarchal females and their male peers. Among the Tsimane, we observe a age-dependent *male* bias in total lymphocyte count between regularly cycling females and their age-matched male counterparts but find no robust sex bias in total lymphocyte count between regularly cycling females in the USA and their male peers. We find a age-dependent male bias in total lymphocyte between pregnant women and their male counterparts in both populations, but this effect is seen across a wider range of ages among the Tsimane (Table 2). There is an age-dependent *female* bias in total lymphocyte count between postmenopausal females and males in the USA, but no evidence of sex differences among the postmenopausal females and males among the Tsimane. -->

<!-- In both populations, we find an age-dependent *female* bias in neutrophil count between regularly cycling females and their male counterparts, but this bias is seen across 8 more years of the reproductive lifespan within the USA compared to the Tsimane. In both populations, we also observe a robust age-dependent female bias in NLR between pregnant females and their age-matched male counterparts, but this effect is larger within the USA. However, the greatest absolute sex difference in NLR among the Tsimane occurs between pregnant females at age `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Age"]` and their age-matched male peers, with a female ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Male Value"]`. As shown in Figure 1C and Figure 1D, the estimated peak NLR among Tsimane women is substantially lower than the estimated peak in NLR among pregnant US women and occurs among women at a later age. As shown in Table 3, the greatest absolute sex difference in NLR within the US population occurs between pregnant females age `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Age"]` and their age-matched male peers, with a female ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Male Value"]`. Similarly, there is an age-dependent male bias in NLR between postmenopausal females and males in the USA, but no evidence of sex differences among the postmenopausal females and males among the Tsimane.  -->

<!-- ##### In the USA, premenarchal females have lower eosinophil and monocyte counts than their male counterparts - a pattern not observed among the Tsimane -->

<!-- Among premenarchal females and their age-matched male peers in the USA, we find evidence of modest-yet-robust `r tolower(get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Sex Bias"])` bias in eosinophil count between ages `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","EOS", "Pre-Menst")[,"Max"]` and monocyte counts between ages `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "NHANES","MON", "Pre-Menst")[,"Max"]` and a modest-yet-robust `r tolower(get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Sex Bias"])` bias in total lymphocyte count from `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Min"]` to `r get.values(table_2, "NHANES","LYM", "Pre-Menst")[,"Max"]` - and again at age `r get.values(table_2, "NHANES","LYM", "Pre-Menst", 2)[,"Min"]` (Figure 1A, Figure 1C, Table 2). Among premenarchal Tsimane females and their age-matched male peers, we find evidence of modest-yet-robust *`r tolower(get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Sex Bias"])`* bias in total lymphocyte count between the ages of `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Min"]` and `r get.values(table_2, "THLHP","LYM", "Pre-Menst")[,"Max"]`, with no evidence of any other strong sex differences (Figure 1B, Figure 1D, Table 2). See Table 3 for the ages and values at which the absolute difference in white blood cell counts and NLR between premenarchal females and their male counterparts are greatest, separated by population and immune marker.  -->

<!-- Within the USA, regularly cycling females have higher neutrophil and white blood cell counts compared to their age-matched male peers between the ages of `r get.values(table_2, "NHANES","NEU", "Cycling")[,"Min"]` and `r get.values(table_2, "NHANES","NEU", "Cycling")[,"Max"]`. We find that regularly cycling females also have higher NLR from ages `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Min"]` to `r get.values(table_2, "NHANES","NLR", "Cycling")[,"Max"]`. Conversely, from ages `r get.values(table_2, "NHANES","MON", "Cycling")[,"Min"]` to `r get.values(table_2, "NHANES","MON", "Cycling")[,"Max"]`, regularly cycling females have lower monocyte counts. For all other ages and immune markers, estimated mean values among cycling females overlap with estimated mean values among age-matched males (Figure 1A, Figure 1C, Table 2). Among the Tsimane, regularly cycling females have lower total lymphocyte counts compared to their age-matched male peers between the ages of `r get.values(table_2, "THLHP","LYM","Cycling")[,"Min"]` and `r get.values(table_2, "THLHP","LYM", "Cycling")[,"Max"]`. For all other ages and immune markers, estimated mean values among cycling females overlap with estimated mean values among age-matched males (Figure 1B, Figure 1D, Table 3).  -->

<!-- ##### Pregnancy is the primary driver of sexual dimorphism in most immune measures within both populations -->

<!-- Among pregnant females in the USA, the estimated non-linear effects of age on neutrophil count `r get.statistics(model_summaries, "USA","NEU","s(Age):Pregnant")` and NLR `r get.statistics(model_summaries, "USA","NLR","s(Age):Pregnant")` are robust (Figure 1A, Figure 1C). Pregnant women at both younger and older ages exhibit relatively lower neutrophil and total white blood counts and NLR (and therefore increased similarity to their age-matched cycling, postpartum, and male peers). We also find negative effects of parity on neutrophil count `r get.statistics(model_summaries, "USA","NEU","s(Parity):Pregnant")`, monocyte count `r get.statistics(model_summaries, "USA","MON","s(Parity):Pregnant")`, and NLR `r get.statistics(model_summaries, "USA","NLR","s(Parity):Pregnant")` among pregnant females in the USA (Figure 2, , Table S4). Controlling for parity, neutrophil count and total white blood cell count among pregnant females are higher than their age-matched male peers from ages `r get.values(table_2, "NHANES","NEU", "Pregnant")[,"Min"]` to `r get.values(table_2, "NHANES","NEU", "Pregnant")[,"Max"]`. Consequently, NLR is higher from ages `r get.values(table_2, "NHANES","NLR", "Pregnant")[,"Min"]` and `r get.values(table_2, "NHANES","NLR", "Pregnant")[,"Max"]`. As shown in Table 3, the greatest absolute sex difference in NLR within the US population occurs between pregnant females age `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Age"]` and their age-matched male peers, with a female ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "NHANES","NLR","Pregnant")[,"Male Value"]`. Conversely, pregnant women in the USA have *lower* total lymphocyte count than their age-matched male peers between ages `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Min"]` and `r get.values(table_2, "NHANES","LYM", "Pregnant")[,"Max"]` and *lower* eosinophil counts than their age-matched male peers between ages `r get.values(table_2, "NHANES","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "NHANES","EOS", "Pregnant")[,"Max"]`. The only immune marker that varies between postpartum women in the USA and their age-matched male peers is monocyte count, which is lower among postpartum females from ages `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Min"]` to `r get.values(table_2, "NHANES","MON", "Postpartum")[,"Max"]`. -->

<!-- Among pregnant Tsimane females, NLR is higher than their age-matched male peers from ages `r get.values(table_2, "THLHP","NLR", "Pregnant")[,"Min"]` and `r get.values(table_2, "THLHP","NLR", "Pregnant")[,"Max"]`, respectively. The greatest absolute sex difference in NLR among the Tsimane occurs between pregnant females at age `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Age"]` and their age-matched male peers, with a female ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "THLHP","NLR","Pregnant")[,"Male Value"]`. As shown in Figure 1C and Figure 1D, the estimated peak NLR among Tsimane women is substantially lower than the estimated peak in NLR among pregnant US women and occurs among women at a later age. Pregnant Tsimane women exhibit *lower* total lymphocyte count than their age-matched male peers between ages `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Min"]` and `r get.values(table_2, "THLHP","LYM", "Pregnant")[,"Max"]` and *lower* eosinophil counts than their age-matched male peers between ages and `r get.values(table_2, "THLHP","EOS", "Pregnant")[,"Min"]` and `r get.values(table_2, "THLHP","EOS", "Pregnant")[,"Max"]`. The only immune marker that varies between postpartum Tsimane women and their age-matched male peers is monocyte count, which is *higher* among postpartum females from ages `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Min"]` to `r get.values(table_2, "THLHP","MON", "Postpartum")[,"Max"]`.  -->

<!-- ##### Among the Tsimane, the transition to menopause is marked by less severe and shorter-lived sex biases in immune status -->
<!-- Among postmenopausal women in the USA and their age-matched male peers, we find a `r tolower(get.values(table_2, "NHANES","LYM", "Menopause")[,"Sex Bias"])` bias in total lymphocyte count between ages `r get.values(table_2, "NHANES","LYM", "Menopause")[,"Min"]` and `r get.values(table_2, "NHANES","LYM", "Menopause")[,"Max"]`. Conversely, we find a `r tolower(get.values(table_2, "NHANES","NEU", "Menopause")[,"Sex Bias"])` bias in neutrophil count from ages `r get.values(table_2, "NHANES","NEU", "Menopause")[,"Min"]` to `r get.values(table_2, "NHANES","NEU", "Menopause")[,"Max"]`, monocyte count from ages `r get.values(table_2, "NHANES","MON", "Menopause")[,"Min"]` to `r get.values(table_2, "NHANES","MON", "Menopause")[,"Max"]`, and NLR from ages `r get.values(table_2, "NHANES","NLR", "Menopause")[,"Min"]` to `r get.values(table_2, "NHANES","NLR", "Menopause")[,"Max"]`. In the USA, the greatest absolute sex difference in NLR among postmenopausal females and their age-matched male counterparts occurs at age `r get.abs.values(table_s3, "NHANES","NLR","Menopause")[,"Age"]`, with a female ratio of `r get.abs.values(table_s3, "NHANES","NLR","Menopause")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "NHANES","NLR","Menopause")[,"Male Value"]`. Among postmenopausal Tsimane women and their age-matched male peers, we find a `r tolower(get.values(table_2, "THLHP","NLR", "Menopause")[,"Sex Bias"])` bias in NLR between ages `r get.values(table_2, "THLHP","NLR", "Menopause")[,"Min"]` and `r get.values(table_2, "THLHP","NLR", "Menopause")[,"Max"]`. Among the Tsimane, the greatest absolute different in NLR among postmenopausal females and their age-matched male peers occurs age `r get.abs.values(table_s3, "THLHP","NLR","Menopause")[,"Age"]`, with a female ratio of `r get.abs.values(table_s3, "THLHP","NLR","Menopause")[,"Female Value"]` and a male ratio of `r get.abs.values(table_s3, "THLHP","NLR","Menopause")[,"Male Value"]`.  -->

### Discussion

This study shows that the direction and magnitude of sex differences in immune status vary by age, and broader ecological context. 

In both populations, sexual dimorphism in NLR (a simple yet robust indicator of immunological homeostasis) is greatest between pregnant females and their age-matched male counterparts, with pregnant females exhibiting substantially higher NLR. Within the USA, we find a robust non-linear effect of age on NLR among pregnant females that we do not observe among the Tsimane. This finding may be due to age-dependent levels of estradiol during pregnancy within the USA, considering estradiol production among females in industrialized populations exhibits a similar non-linear relationship with age, peaking around the age of 30 [@lephart_naftolin21]. Among females in the USA, becoming pregnant at earlier or later ages may be characterized by lower levels of estradiol due to age-dependent ovarian function, while chronic demands on energy allocation among Tsimane women may result in less variability in hormone production by age. Considering pregnancy is associated with short-term reduction of ovarian hormone production [@arslan_etal06; @barrett_etal14], the absence of a non-linear effect of age on NLR among pregnant Tsimane women may also be due to earlier age at first birth and shorter interbirth intervals. 

**Pregnancy = when female disease risk is more similar to men (e.g., increased viral susceptibility, reduced autoimmune risk), but when immune cell counts are most different!**

Among pregnant females in the USA, we also observe a negative effect of parity on neutrophil count and NLR, with high parity associated with reduced male-female differences and reduced within-sex differences between pregnant, regularly cycling, and postpartum females. Estimated NLR in the USA is highest among primiparous pregnant females around 29 years of age, with estimated values above pre-established NLR thresholds associated with "mild to moderate inflammation" [@zahorec21]. Approximately ~84% of females in the USA have at least one child [@martinez_daniels23], the average number of live births per woman in 2023 was 1.78. According to 2021 data from the Centers for Disease Control, the average woman in the USA gives birth for the first time at ~27 years old. National trends in the USA are therefore remarkably close to the demographic that we find to be the most "at risk" for particularly high NLR during pregnancy, based on our modeling. These patterns may explain pre-existing evidence that autoimmune diseases flare and/or emerge after pregnancy [@khashan_etal11] and are predominantly diagnosed before the average age of menopause in industrialized populations [@angum_etal20], while cancer risk is also elevated after birth and cancers diagnosed post-delivery have significantly poorer prognosis [@lefrere_etal21; @zhang_etal22]. During pregnancy, immediate deleterious effects of a big spike in NLR may be mitigated by the presence of the fetus/placenta and the associated suite of mechanisms known to induce fetal tolerance (e.g., regulatory T cell proliferation) [@Erlebacher2013a]. Deleterious effects on the mother may be more likely to emerge after delivery, when fetal/placental cues are removed. It is possible that the comparatively acute spike in NLR during pregnancy among women in the USA, in addition to relatively low rates of extended on-demand breastfeeding [@flaherman_etal11], may result in incomplete or slower postpartum immunological recovery and thus pave the way for immune dysregulation. Parsing out the effects of month since delivery on immune markers (and comparing them to the effects of month of pregnancy) will likely illuminate effects we were not able to separate out in this study, considering all women who had given birth within the last 12 months were binned as "postpartum". 

Of the ~16% of females in the USA who do not give birth, it is unclear how many suffer from a fertility issue (e.g., endometriosis) and how many simply choose to forgo childbearing in the absence of any underlying medical issues. Likewise, it is unclear whether or not the risk of autoimmune disease, infection, and/or cancer varies between nulliparous females with and without underlying fertility issues that may separately impact disease risk. Given that we find comparatively little sexual dimorphism between regularly cycling females and their age-matched counterparts (even in the USA) and only minor effects of parity on immune status among postmenopausal females in the USA, we pose the following questions. Do nulliparous females without any underlying fertility issues have a reduced chance of developing autoimmune disease compared to low-parity parous women who begin their reproductive careers in the late-twenties/early-thirties? And if so, how does the risk within this category compare to that among high-parity women who start their reproductive career relatively early? 

The differences we find between premenarchal females in the USA and their age-matched male peers aligns with prior research on industrialized populations indicating an early male bias in eosinophil [@hartl_etal20] and monocyte counts [@patel_yona19]. The absence of any such early male bias in eosinophil and monocyte count among the Tsimane, however, suggests that this pattern is not universal.  We find that eosinophil and monocyte counts do not vary by age or reproductive state among all females in the USA, suggesting that these immune cell types may be less responsive to changes in ovarian sex hormone production than neutrophils and lymphocytes - even after puberty. Since these population differences cannot be attributed to differences in parity and appear less likely to stem from hormonal differences, more research is needed to elucidate the underlying mechanism(s). Perhaps these population differences stem from the relative absence of exposure to infectious disease in the USA, which may expose the effects of differential gene expression on baseline immune status beginning in childhood. 

<!-- Ex vivo studies have shown that male monocytes produce more inflammatory cytokines than females when stimulated with LPS (Bouman et al 2004) but that this effect does not appear to be due to female sex hormones (Bouman et al). -->

**Are these unanswerable in the epi literature? Do they run counter to existing narratives, hence flagging these as outstanding questions? Is this the part where you say what results imply in terms of autoimmune risk?**

Lastly, our results indicate that the transition to menopause within the US population is marked by a comparatively greater and more sustained reduction in NLR among females and correspondingly larger and longer-lasting sex differences in this functional age group. This suggests that the transition to menopause is characterized by a more severe drop-off in ovarian hormone production among females in the USA and/or greater sensitivity of the immune system to the hormonal changes that occur during menopause. Taken together, our findings indicate that females in the USA (especially those who have one or more pregnancies) experience a much higher degree of overall variability in neutrophil count and NLR across the lifespan when compared to their Tsimane counterparts, presumably as a result of greater vacillations in ovarian sex hormone production during different reproductive states.

**The PCH and hormone hypothesis described in intro are not directly addressed in the Discussion. What can be said about them here, and next steps?**

<!-- It is also plausible that women in the USA may exhibit greater neutrophil expansion and higher NLR during pregnancy due to comparatively greater exposure to fetal antigens during gestation. Maternal weight influences the development of both the placenta and the fetus, with higher maternal weight strongly associated with fetal overgrowth [@stang_huffman16], increased nutrient transport across the maternal-placental interface [@acosta_etal15; @jansson_etal13; @jones_etal10; @rosario_etal12], higher rate of trophoblast apoptosis, and increased levels of cell-free fetal DNA found in maternal circulation [@vora_etal09]. These patterns have been mostly studied within industrialized populations (and in animal models), but there is a possibility that these same mechanisms may produce differences between Tsimane and USA women who have similar BMI but who are experiencing different energetic throughput due to a variety of ecological factors (e.g., concurrent infection).  -->

<!-- Our results indicate that overall *intrapopulation* sex differences in leukocyte differential and NLR during the reproductive years primarily reflect immunological changes that occur during pregnancy. Our findings also indicate that overall *interpopulation* differences in the magnitude of sexual dimorphism during this part of the lifespan reflect comparatively greater increase in neutrophil and monocytes counts and NLR during pregnancy among women in the USA and comparatively greater decrease in total lymphocyte and eosinophil count during pregnancy among Tsimane women - echoing prior research [@Hove2020]. Additionally, our findings suggest population-specific effects of menopause on sexual dimorphism in leukocyte differential and NLR. Among women in the USA, the transition to menopause is associated with significantly *reduced* total leukocyte, neutrophil, and monocyte counts and NLR. In contrast, NLR is the only outcome measure for which we found a robust sex bias among postmenopausal Tsimane women and their male counterparts. These patterns suggest that the drop-off in hormone production following menopause exerts stronger corresponding downstream effects on immune markers among women in the USA. -->

#### Limitations
In this study, we were limited to data on immune cell counts. While useful in establishing a general understanding of immune profile, variability in cell counts does not perfectly map on to variation in cellular function or underlying gene expression. Future comparative work is needed to further investigate population-level differences in sexual dimorphism at deeper functional levels (e.g., oxidative bursts and release of extracellular traps by neutrophils). We especially recommend that future research parse out the effects of age, sex, female reproductive status, parity, and study population on neutrophil subtypes, especially low-density neutrophils (LNDs), as these have been identified as key players in the etiology of autoimmune disease [@ning_etal22]. 

Additionally, the comparisons we draw between the USA and the Tsimane should be interpreted in context. There is significant variation in socio-ecological conditions *within* non-industrialized societies that must be carefully considered, including differences in physical environment (e.g., altitude, seasonality, type of pathogen exposure) and social structure (e.g., population density, matriarchal versus patriarchal customs). For example, the Shuar (another indigenous population living in the Amazon) have distinct pathogen exposure profiles compared to the Tsimane [@Blackwell2011c]. Furthermore, there is considerable socio-ecological variation within industrialized societies (e.g., immigration status, ethnic background) that we chose to collapse based on the aims of this particular study. 

#### Conclusions
The results of this study support our overarching hypothesis that the magnitude of sexual dimorphism in leukocyte differential and neutrophil-to-lymphocyte ratio is comparatively attenuated among the Tsimane. Specifically, we find outsized effects of pregnancy and menopause on female immune status in the USA and therefore outsized effects of these reproductive phases on interpopulation variation in the degree and duration of overall sexual dimorphism. We find that a strong negative effect of parity on neutrophil count, total white blood cell count, and NLR among pregnant females in the USA results in reduced sexual dimorphism between high-parity pregnant females and their age-matched male peers in this population. Conversely, we do not find a strong effect of parity on immune status among Tsimane females, regardless of reproductive state. Based on these findings, we argue that present understanding of “normal” sexual dimorphism in immune function is highly specific to the post-industrialized contexts in which these patterns have predominantly been studied and is heavily influenced by the inclusion/exclusion criteria used by existing studies (e.g., exclusion of pregnant females). We strongly recommend that any study of sex differences in immune status and disease risk explicitly consider the age, reproductive state, and (when applicable) parity of all females included in the study sample. 

\newpage

### Methods

#### The Tsimane
The Tsimane are a relatively small subsistence population inhabiting the Bolivian Amazonian River basin. Among the Tsimane, chronic exposure to diverse pathogens causes high infectious disease morbidity and mortality across all ages [@gurven_etal19; @kaplan_etal15; @martin_etal13], while incidence of allergies, atopy, autoimmune disease, obesity, and atherosclerosis is low [@gurven_etal07; @Gurven2008a; @kaplan_etal17]. As a result of elevated pathogen burden, Tsimane individuals exhibit high levels of immune activation compared to Western clinical standards [@Blackwell2016; @Blackwell2011c]. Such high investment in immune function results in trade-offs with growth during development [@Foster2005a] and high resting metabolic rate during adulthood [@gurven_etal16], reflecting the substantial energetic demands of coping with pathogenic threats. The Tsimane are also a natural fertility population, with limited access to contraception and breastfeeding alternatives [@trumble_etal23a]. As a result, Tsimane women have an average 9 live births over the reproductive lifespan [@McAllister2012a] and exhibit nearly ubiquitous rates of on-demand breastfeeding following parturition, with a mean infant weaning age of 27 months [@Martin2016a]. 

#### Datasets
We used cross-sectional and longitudinal clinical and demographic data collected by the Tsimane Health and Life History Project (http://tsimane.anth.ucsb.edu/index.html) [@Blackwell2011c; @Blackwell2015c; @Blackwell2016; @gurven_etal17; @martin_etal13] between 2004 and 2014. Approval by the Gran Consejo Tsimane and by institutional review boards at the University of California, Santa Barbara and the University of New Mexico was obtained before any data were collected. Informed consent was provided by participants during a community-wide meeting open to all Tsimane residents and again at the individual level before each medical visit and interview. In the case of minors, parental consent was given before data were collected. Total leukocyte count was obtained via venous blood draws and determined with a QBC Autoread Plus dry hematology system (QBC Diagnostics). Relative fractions of neutrophils, eosinophils, lymphocytes, and monocytes were then measured manually by microscopy with a hemocytometer. Menarche was based on self-reported presence/absence of first menstrual cycle. Pregnancy status was determined during medical visits based on date of last menses, with urinary pregnancy tests administered by the physician when pregnancy was suspected. Pregnancies were cross-validated against subsequent annual demographic and census interviews, allowing detection of pregnancies that occurred between medical visits and pregnancies that went undetected during previous physician examinations [@Blackwell2015c]. Menopause was determined based on reported absence of a menstrual cycle over the past 6 months.

To obtain a representative sample from the United States, we used publicly available cross-sectional NHANES data collected by the Centers for Disease Control (https://www.cdc.gov/nchs/nhanes/index.htm) between 2003 and 2016. Total and differential leukocyte counts were measured using the Coulter method. Females who were under the age of 8 (for which reproductive data were redacted) or who specifically reported absence of menarche were binned as premenarchal. Females who were not currently pregnant, had not given birth within the preceding 12 months, and who reported a regular menstrual cycle either at time of exam or within the preceding two months were binned as regularly cycling. Females who self-reported being pregnant and/or had a positive urine test at the time of exam were categorized as currently pregnant. Women who were not currently pregnant but had given birth within the past 12 months were considered postpartum. Finally, females who reported absence of regular menstruation due to menopause were binned as postmenopausal.

#### Sample Selection
Final THLHP and NHANES data sets were limited to males and females ages 2 to 84 with recorded white blood cell differential and body mass index (BMI). NLR was calculated by dividing neutrophil count by total lymphocyte count. We excluded `r early_meno$n` females from the NHANES dataset who reported reaching menopause before age 40 and `r late_meno$n` individual who reported regular menstruation after the age of 70 (effectively removing outliers associated with very early versus delayed menopause and/or errors in the data). We also removed extreme values for immune cell counts and NLR by limiting our sample to values which fell between population-pooled 1% and 99% percentiles. This exclusion criterion resulted in `r sum(removed_data$removed)` data points (`r str_c(percent = format(round(sum(removed_data$removed)/sum(removed_data$original)*100, digits = 2), nsmall = 2), "%")`) being removed out of the initial pooled dataset consisting of `r sum(removed_data$original)` unique values. Given the ubiquity of chronic parasitic infections among the Tsimane, we did not exclude individuals with known infection or illness. In both populations, BMI varies considerably by age and sex (Figure S1), so we z-scored all BMI values using sex and age-specific mean values. No exclusions were based on medical diagnoses. Finally, we used the *matchit* package [@ho_etal11] to create matched THLHP and NHANES samples stratified by age, sex, and female reproductive phase (specifying nearest neighbor matching on propensity score) (Figure S2, Table S1). 

#### Statistical Analyses

<!-- model_formula <- as.formula("as.integer(value) ~ RepPhase + s(Age, by = RepPhase) + BMI_zscore + s(NumPartos, by = RepPhase) + s(id, bs = 're')") -->

All models were executed in R 4.3.2 (https://cran.r-project.org) using the *mgcv* package [@wood11]. We employed generalized additive models (GAMs) to estimate the non-linear population-specific effects of age by sex and female reproductive phase (premenarchal, regularly cycling, pregnant, postpartum, postmenopausal) on total leukocyte,  neutrophil, total lymphocyte, eosinophil, and monocyte count and neutrophil-to-lymphocyte ratio (NLR). All models also accounted for the fixed effects of z-scored BMI and the smoothed effects of parity, stratified by female reproductive phase (males and premenarchal females were all assigned parity values of zero). Due to repeat sampling of individuals within the THLHP dataset (Table S2), all THLHP-specific models accounted for the group-level random effects of participant identification number. 

\newpage

### Figures
<!-- Provide all images at final size. While figures may be sized conservatively to save page space, PNAS reserves the right to make the final decision on figure size in published articles and authors may be asked to shorten manuscripts that exceed the stated length requirement. -->
<!-- Small: approximately 9 cm x 6 cm -->
<!-- Medium: approximately 11 cm x 11 cm -->
<!-- Large: approximately 18 cm x 22 cm -->
<!-- Ensure that all numbers, letters, and symbols are no smaller than 6 points (2 mm) and no larger than 12 points (6 mm) after reduction. Keep text sizing consistent within each graphic. -->
<!-- Preassemble all composite figures.  -->
<!-- Submit images in these file formats: TIFF, EPS, PDF, or PPT. -->

<!-- Include figure legends immediately after referencing the figure in the manuscript. Ensure that figure legends adhere to the following guidelines:  -->

<!-- For figures with multiple panels, the first sentence of the legend should be a brief overview of the entire figure. Explicitly reference and describe each panel at least once in the figure legend. -->
<!-- Include clearly labeled error bars in all graphs and describe them in the figure legend. -->
<!-- State whether a number that follows the ± sign is a standard error (SEM) or a standard deviation (SD). -->
<!-- When applicable, provide the P value, magnification, or scale bar information.  -->
<!-- Indicate the number of independent data points (N) represented in a graph in the legend. -->
<!-- Ensure that numerical axes on all graphs go to 0, except for log axes. -->

<!-- Combination halftones: Common examples are color or grayscale figures containing halftone and line art -->
<!-- elements. The preferred resolution for this type of image is between 600 and 900 ppi at publication size.  -->

![Fig 1. Non-linear effects of age by sex and female reproductive phase on (A) leukocyte differential among the Tsimane, (B) leukocyte differential in the USA, (C) NLR among the Tsimane, and (D) NLR in the USA (Table 1). Estimated values are standardized by parity (set to 0 live births for premenarchal females and 3 live births for regularly cycling, pregnant, postpartum, and postmenopausal females) and z-scored BMI (set to 0). Solid lines correspond to estimated mean value, shaded regions correspond to 95% credible intervals. Please note that each variable is plotted on its own scale.](../output/figure_1.png){width=100%}

\newpage

![Fig 2. Non-linear effects of parity by female reproductive phase on (A) leukocyte differential among the Tsimane, (B) leukocyte differential in the USA, (C) NLR among the Tsimane, and (D) NLR in the USA (Table 1). Estimated values are standardized by age (set to 24 years for regularly cycling, pregnant, and postpartum females and set to 65 years for postmenopausal females) and z-scored BMI (set to 0). Solid lines correspond to estimated mean value, shaded regions correspond to 95% credible intervals. Please note that each variable is plotted on its own scale.](../output/figure_2.png){width=80%}

\newpage

### Tables
<!-- Tables -->
<!-- Ensure that the table is in an editable Word, RTF, or LaTeX format. -->
<!-- Include a brief title (above) and footnotes (below) the table. -->
<!-- Avoid multipart tables (Table 1A, Table 1B). -->

```{r, echo=F}
knitr::kable(table_1, 
             caption = "Table 1. Description of immune markers used in this study.")
```

\newpage

```{r, echo=F}
knitr::kable(table_2 %>% 
               mutate(`Reproductive Phase` = case_when(`Reproductive Phase` == "Pre-Menst" ~ "Premenarchal",
                                           `Reproductive Phase` == "Menopause" ~ "Postmenopausal",
                                           TRUE ~ `Reproductive Phase`),
                      `Reproductive Phase` = ordered(`Reproductive Phase`,
                            levels = c("Premenarchal","Cycling","Pregnant",
                                       "Postpartum","Postmenopausal"))) %>%
               select(-Group) %>%  
               arrange(`Reproductive Phase`, Population,Measure) %>% 
               select(-`Age Span`, -`% Age Span`), 
             caption = "Table 2. Ages at which the 95% credible intervals for estimated values between males and females do not overlap, separated by population, measure, and female reproductive phase. Estimated values presented in this table are standardized by parity (set to 0 live births for premenarchal females and 3 live births for regularly cycling, pregnant, postpartum, and postmenopausal females) and z-scored BMI (set to 0).")
```

\newpage

### Data availability
R code for the statistical analyses are published on a public GitHub repository (https://github.com/carmenhove/sex_differences). Individual-level data are stored in the THLHP Data Repository and are available through restricted access for ethical reasons. THLHP's highest priority is the safeguarding of human subjects and minimization of risk to study participants. The THLHP adheres to the "CARE Principles for Indigenous Data Governance" (Collective Benefit, Authority to Control, Responsibility, and Ethics), which assure that the Tsimane (i) have sovereignty over how data are shared, (ii) are the primary gatekeepers determining ethical use, (iii) are actively engaged in the data generation, and (iv) derive benefit from data generated and shared for use whenever possible. The THLHP is also committed to the "FAIR Guiding Principles for scientific data management and stewardship" (Findable, Accessible, Interoperable, Reusable). Requests for individual-level data should take the form of an application that details the exact uses of the data and the research questions to be addressed, procedures that will be used for data security and individual privacy, potential benefits to the study communities, and procedures for assessing and minimizing stigmatizing interpretations of the research results (see the following webpage for links to the data sharing policy and data request forms: https://tsimane.anth.ucsb.edu/data.html). Requests for individual-level data will require institutional IRB approval (even if exempt) and will be reviewed by an Advisory Council composed of Tsimane community leaders, community members, Bolivian scientists, and the THLHP leadership. The study authors and the THLHP leadership are committed to open science and are available to assist interested investigators in preparing data access requests.

### Acknowledgements
We offer our profound thanks to all participants who participated in this study, the Tsimane Gran Consejo, and the THLHP mobile team and staff.

### Funding
The THLHP was supported by the National Institute of Health/National Institute on Aging (R01AG054442, R01AG024119, R56AG024119, P01AG022500) and National Science Foundation (BCS0422690). J.S. acknowledges Institute for Advanced Study in Toulouse funding from the French National Research Agency (ANR). 

\newpage

### References
<div id="refs"></div>

<!-- Some studies have found that women possess higher neutrophil and monocyte counts and higher neutrophil-lymphocyte ratio than men [@chen_etal16a; @li_etal20; @bain_england75], while others have reported no difference [@thammawong_etal] or a male bias in these measures [@hu_etal20; @pellegrino_etal23; @patel_yona19]. Other studies have found a male bias in eosinophil [@hartl_etal20] and monocyte counts. Higher neutrophils and NLR during infection during childhood [@casimir_etal10]. Full blood counts for lymphocytes were higher in male and granulocytes higher in female [@abdullah_etal12]. Prior research on sex differences in leukocyte differential and NLR are convoluted, presumably due to varying inclusion criteria and age groups studied. -->

<!-- some evidence that sex biases are either attenuated or reversed in later ages [@chen_etal16a] -->

<!-- It is also possible that the number of pregnancies a woman experiences over the lifespan may further moderate her immune system and corresponding disease risk. The Pregnancy Compensation Hypothesis (PCH) proposes that low parity, a phenomenon most commonly observed among females in post-industrial societies, may amplify evolved sex differences in immune function by creating a mismatch between proposed 'compensatory' immunological responses (evolved to induce fetal tolerance) and less time spent gestating over the lifespan [@Natri2019]. It is also possible that the effects of parity may vary across ecologies, due to different hormonal response patterns during pregnancy. For example, women in industrialized contexts exhibit higher progesterone levels during the peri-ovulatory and peri-implantation period [@vitzthum_etal04], suggesting that the hormonal changes induced by pregnancy may be more extreme in contexts where energetic constraints are relatively relaxed. To date, however, regular exclusion of pregnant and/or menopausal women and binning of participants into categorical age groups [@bain_england75; @hu_etal20; @li_etal20; @pellegrino_etal23; @thammawong_etal] precludes a clear picture of how age, female reproductive status, and parity combine to impact sexual dimorphism in underlying immune status. -->

<!-- estradiol, estrone, progesterone, and testosterone are all elevated above baseline during pregnancy [@schock_etal16], while immune changes during gestation may drive a large proportion of sexual dimorphism within each population. Given the reversal of pregnancy-induced hormonal patterns, suppression of ovulation [@neville_neifert83] and the removal of the need for fetal tolerance, postpartum females may exhibit less divergent immune profiles from their male counterparts. Given the drop-off in female hormone production after menopause and some evidence that sex biases are either attenuated or reversed in later ages [@chen_etal16a] we expect that sexual dimorphism in immune status will be reduced among postmenopausal women and age-matched men. Even in otherwise heavily studied populations, the effects of female reproductive phase are relatively understudied.  -->

<!-- Average age of natural menopause is around 50 years of age [@gold11; @zamaniyan_etal20].  -->

<!-- Some studies show an early but short-lived male bias in infant mortality [@aghai_etal20, while others have shown a relatively higher mortality rate among females [@almossawi_etal21].  -->

<!-- Most usual onset for Sjögren syndrome is middle age (in the fifties) [@garcia-carrasco_etal02], systemic lupus erythematosus is between 20 and 40 years of age [@ohta_etal13], hashimoto thyroiditis occurs between women age 30 to 50 [@mincer_jialal24] (and risk is particularly elevated after pregnancy), sclerodema at age 50 [@moinzadeh_etal20], myasthenia gravis at mid life and beyond [@aarli99]. -->

<!-- * "However, human male infants have higher monocyte and basophil counts compared to females up until 13 months of age, at least in high pathogen burden settings. NK cell frequencies are higher in male than female children. Infant males have greater pro-inflammatory responses than females following stimulation with either LPS or mitogens, all suggesting that males develop more robust innate immunity compared with females in early life." [@Klein2016].  -->
